---
title: "Leverage and Influence: Which Gaugings Are Steering Your Fit"
author: "Jonathan Payne, Environment Agency Flood Forecasting"
date: "September 2026"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Leverage and Influence: Which Gaugings Are Steering Your Fit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE, echo = TRUE, collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5
)
library(reach.rate)
library(data.table)
library(ggplot2)
```

Here's a question `plot_rating_residuals()` can't answer: if one particular
gauging had never been taken -- lost in the field, a bad day, whatever --
would your fitted `C`, `a`, and `n` look basically the same, or would they
land somewhere noticeably different? A gauging can sit right on its own
limb's curve, with a small, unremarkable residual, and *still* be the
reason the curve is where it is. This vignette is about spotting that
gauging before it surprises you, using `flag_influential_gaugings()` and
`plot_rating_leverage()` -- and about explaining, in plain language, what
the numbers those functions produce actually mean.

# Two questions that sound similar but aren't

- **"Does this gauging fit well?"** -- a residual question. `plot_rating_residuals()`
  answers it: observed minus fitted discharge, for every gauging. A point
  right on the curve has a small residual.
- **"Did this gauging decide where the curve goes?"** -- an influence
  question. A gauging can have a *tiny* residual precisely because the
  fit bent itself to reach it, especially if that gauging sits alone at
  one end of a limb with nothing nearby to hold the curve in place. That's
  what this vignette covers.

Think of fitting a curve like balancing a plank across a few scattered
weights. A weight near the middle, surrounded by others, barely moves the
plank if you nudge it -- there's support all around. A single weight way
out at one end, with nothing else nearby, can tip the whole plank on its
own. Whether that weight looks "wrong" once the plank has settled (its
residual) is a completely different question from how much *leverage* it
had to begin with.

# A worked example

Twenty-six ordinary gaugings across a range, plus one that sits alone at
the sparse top end and reads about 25% higher than the smooth curve
running through everything else would predict -- not an absurd value,
just an isolated one, at the one place nothing else backs it up.

```{r worked-data}
set.seed(1)
stage_m <- seq(0.3, 3, by = 0.1)
discharge_cms <- 5 * stage_m^1.5 * exp(rnorm(length(stage_m), sd = 0.03))
# one isolated, high gauging at the sparse top end
stage_m <- c(stage_m, 3.5)
discharge_cms <- c(discharge_cms, 5 * 3.5^1.5 * 1.25)

fit <- rate_optimise(discharge_cms, stage_m, n_bounds = c(1.3, 1.7))
fit@limbs[, .(C, a, n, rmse_cms, r_squared)]
```

`r_squared` looks fine, and `vignette("rating_curves_guide")`'s usual
residual check wouldn't obviously single this gauging out either. Ask the
influence question instead:

```{r flag}
flagged <- flag_influential_gaugings(fit)
flagged@gaugings[order(-cooks_distance)][1:5, .(stage_m, discharge_cms, residual_cms, leverage, cooks_distance, influential)]
```

The top row is exactly the gauging planted above -- flagged `influential`,
with a Cook's distance an order of magnitude past everything else in the
table. `plot_rating_leverage()` shows the same thing across every limb at
once, leverage against Cook's distance, flagged points picked out in red:

```{r plot-leverage, fig.alt="Leverage against Cook's distance for every gauging, one facet per limb, with influential points highlighted in red"}
plot_rating_leverage(fit)
```

# What "leverage" and "Cook's distance" actually mean

**Leverage** is about *position only* -- how alone a gauging is, not
whether its value looks right. A gauging in the crowded middle of a limb
has low leverage: there are others nearby, so no single point can swing
the curve far. A gauging isolated at the sparse end of a limb has high
leverage just by being out there, regardless of what discharge it
recorded.

**Cook's distance** combines that with the residual: leverage tells you a
gauging *could* swing the fit; Cook's distance estimates how much it
*actually did*, given both its position and how far its value sits from
what the rest of the data implies. A high-leverage gauging that happens
to land right where the curve would have gone anyway scores a low Cook's
distance -- no harm done. It's the combination of *alone* and *off* that
matters, not either on its own.

Concretely, the plot above and the table before it show this gauging with
leverage around 0.6-0.7 (out of a maximum of 1) against everything else
sitting under 0.15 -- it is, numerically, in a class of its own before its
value is even considered.

## Seeing it directly: refit without the flagged gauging

The most direct way to check what a flagged gauging is doing is the most
literal one: take it out, refit, and look at both curves together.

```{r refit-comparison, fig.alt="Two fitted rating curves overlaid, one using all gaugings and one with the flagged gauging removed, diverging visibly near the top of the range"}
top_idx <- which.max(flagged@gaugings$cooks_distance)
fit_without <- rate_optimise(
  discharge_cms[-top_idx], stage_m[-top_idx], n_bounds = c(1.3, 1.7)
)

curve_dt <- function(f, label) {
  h <- seq(f@limbs$lower_stage_m, f@limbs$upper_stage_m, length.out = 200)
  data.table(stage_m = h, discharge_cms = f@limbs$C * (h - f@limbs$a)^f@limbs$n, fit = label)
}
curves_dt <- rbind(
  curve_dt(fit, "All gaugings"),
  curve_dt(fit_without, "Flagged gauging removed")
)

ggplot() +
  geom_line(data = curves_dt, aes(discharge_cms, stage_m, colour = fit), linewidth = 1.1) +
  geom_point(data = flagged@gaugings, aes(discharge_cms, stage_m, shape = influential), size = 2, colour = "grey30") +
  labs(
    title = "One gauging, two fitted curves",
    x = "Discharge (m^3/s)", y = "Stage (m)", colour = NULL, shape = "Influential"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 13))
```

```{r refit-pct}
pct_shift <- 100 * (fit@limbs$C - fit_without@limbs$C) / fit_without@limbs$C
sprintf("C shifts by %.0f%% depending on whether that one gauging is included", pct_shift)
```

That's the plain-English payoff: one gauging, on its own, moves the
fitted coefficient by a number worth caring about. `flag_influential_gaugings()`
found this without anyone having to refit anything by hand -- the whole
point of the diagnostic is knowing this *before* you've committed to a
curve, not after a second gauging campaign disagrees with it.

# So a gauging is flagged -- now what?

Being flagged is a prompt to look, not a verdict. Three honest
possibilities, and only one of them means "consider removing it":

1. **It's an error.** A transcription slip, a misread staff gauge, a
   units mix-up. Once found, fix it at the source and refit -- this is
   the case flagging is most valuable for, since a small data error can
   otherwise hide behind a perfectly reasonable-looking `r_squared`.
2. **It's real, and it's the only evidence you have out there.** A
   station's single highest-flow gauging, taken during the one flood
   worth capturing, is *supposed* to have high leverage -- there's
   nothing else out there to share the load with. Removing it doesn't
   make the curve more correct at high flow; it just removes the only
   information you had about it. The honest response here is usually
   "keep it, and treat that limb's high end as under-supported" --
   exactly what `flag_extrapolated_limbs()` already separately checks.
3. **It's real, but the rest of the limb is oddly thin nearby.** Nothing
   wrong with the gauging itself; the real issue is there's a gap in
   gauging coverage that happens to hand this one point too much sway.
   The fix is more gaugings in that gap, not fewer at the flagged end.

Never drop a flagged gauging reflexively to improve `r_squared` --
that's precisely the over-refinement NEMS warns against
(`vignette("rating_curves_guide")`'s "Guidance: reading a fit, in order"
section already covers this caution in general; influence is the
specific mechanism it's warning about).

# Where this sits among the other diagnostics

| Question | Function |
|---|---|
| How far off is each point, after fitting? | `plot_rating_residuals()` |
| How much did each point shape the fit, regardless of its residual? | `flag_influential_gaugings()`, `plot_rating_leverage()` |
| Does a limb reach far enough beyond its own gaugings? | `flag_extrapolated_limbs()` |
| How precise are the fitted coefficients overall? | `rate_optimise()`'s `C_se_asymp`/bootstrap columns |

They answer different questions and are meant to be read together, not
as a replacement for one another -- a limb can pass every residual check
and still be quietly propped up by one gauging.

# References

Belsley, D. A., Kuh, E., & Welsch, R. E. (1980). *Regression
Diagnostics: Identifying Influential Data and Sources of Collinearity*.
Wiley.

Cook, R. D. (1977). Detection of influential observations in linear
regression. *Technometrics*, 19(1), 15-18.

St. Laurent, R. T., & Cook, R. D. (1992). Leverage and superleverage in
nonlinear regression. *Journal of the American Statistical Association*,
87(420), 985-990.

# Where to go next

- `vignette("rating_curves_guide")` -- "Diagnosing a fit" for the full
  set of diagnostics this one complements, and "Guidance: reading a fit,
  in order" for the over-refinement caution referenced above.
- `?flag_influential_gaugings`, `?plot_rating_leverage` -- full argument
  reference, including how the default `influential` cutoffs are chosen.
