---
title: "Choosing n_bounds: A Cross-Section Guide"
author: "Jonathan Payne, Environment Agency Flood Forecasting"
date: "August 2026"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Choosing n_bounds: A Cross-Section Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE, echo = TRUE, collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5
)
library(reach.rate)
library(data.table)
library(ggplot2)
library(gridExtra)
library(grid)
```

`vignette("non_standard_optimisation")` covers `n_bounds` mechanically:
what argument to pass, and a before/after comparison of the fit it
produces. It doesn't say how to arrive at the numbers you pass it. This
vignette is the companion piece -- given a channel, or a cross-section
survey of one, what exponent should you expect, and what does that look
like. Read `non_standard_optimisation.Rmd` for the API; read this one for
where its `c(1.4, 1.6)` actually comes from.

# The three canonical controls

`vignette("rating_curves_guide")`'s "Why a power law" section states the
theory in one paragraph: a rectangular weir or channel control gives an
exponent around 1.5; a V-notch or triangular control, around 2.5;
Manning's equation for a wide channel, around 5/3 (Herschy 2009; Rantz et
al. 1982). What those three controls actually look like, side by side:

```{r xs-helpers}
bed_colour <- "#8d6e63"
water_colour <- "#0288d1"
line_colour <- "#4e342e"

# Clip a bed polyline to the part at or below `level`, inserting an exact
# crossing point at every wall/bank so the wetted polygon closes cleanly
# even where the bed has a vertical segment (a repeated x value, which
# stats::approx() can't interpolate through).
clip_below <- function(bed_dt, level) {
  n <- nrow(bed_dt)
  pts <- vector("list", 0)
  wet <- bed_dt$elevation_m <= level
  if (wet[1]) pts[[length(pts) + 1]] <- c(bed_dt$distance_m[1], bed_dt$elevation_m[1])
  for (i in seq_len(n - 1)) {
    x1 <- bed_dt$distance_m[i]; y1 <- bed_dt$elevation_m[i]
    x2 <- bed_dt$distance_m[i + 1]; y2 <- bed_dt$elevation_m[i + 1]
    if (wet[i] != wet[i + 1]) {
      xc <- if (x1 == x2) x1 else x1 + (level - y1) / (y2 - y1) * (x2 - x1)
      pts[[length(pts) + 1]] <- c(xc, level)
    }
    if (wet[i + 1]) pts[[length(pts) + 1]] <- c(x2, y2)
  }
  setNames(as.data.table(do.call(rbind, pts)), c("distance_m", "elevation_m"))
}

# An idealised cross-section: bed outline, a shaded wetted area up to one
# water level, and a caption naming the control and its theoretical n.
xs_diagram <- function(bed_dt, water_level, title, n_label, note, baseline_drop = 0.4) {
  baseline <- min(bed_dt$elevation_m) - baseline_drop
  wet_dt <- clip_below(bed_dt, water_level)
  wet_poly <- rbindlist(list(
    wet_dt,
    data.table(distance_m = rev(wet_dt$distance_m), elevation_m = water_level)
  ))
  ground_poly <- data.table(
    distance_m = c(bed_dt$distance_m, rev(bed_dt$distance_m)),
    elevation_m = c(bed_dt$elevation_m, rep(baseline, nrow(bed_dt)))
  )

  ggplot() +
    geom_polygon(data = ground_poly, aes(distance_m, elevation_m), fill = bed_colour, colour = NA) +
    geom_polygon(data = wet_poly, aes(distance_m, elevation_m), fill = water_colour, alpha = 0.55) +
    geom_line(data = bed_dt, aes(distance_m, elevation_m), colour = line_colour, linewidth = 1.1) +
    geom_hline(yintercept = water_level, colour = water_colour, linetype = "dashed", linewidth = 0.6) +
    labs(title = title, subtitle = bquote(italic(n) ~ "~" ~ .(n_label)), caption = note, x = NULL, y = NULL) +
    coord_cartesian(ylim = c(baseline, max(bed_dt$elevation_m) + 0.3)) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", size = 12),
      plot.subtitle = element_text(size = 11),
      plot.caption = element_text(size = 8, colour = "grey40", hjust = 0),
      axis.text = element_blank(), axis.ticks = element_blank(),
      panel.grid = element_blank()
    )
}
```

```{r xs-three, fig.width=9.5, fig.height=3.6}
rect_bed <- data.table(distance_m = c(-4, -4, 4, 4), elevation_m = c(3, 0, 0, 3))
p_rect <- xs_diagram(rect_bed, water_level = 1.2, "Rectangular weir / channel",
                      "1.5", "Top width constant as head rises")

vnotch_bed <- data.table(distance_m = c(-2.2, 0, 2.2), elevation_m = c(2.0, 0, 2.0))
p_vnotch <- xs_diagram(vnotch_bed, water_level = 1.1, "V-notch / triangular",
                        "2.5", "Top width grows linearly with head", baseline_drop = 0.35)

wide_bed <- data.table(distance_m = c(-8, -5, 5, 8), elevation_m = c(2.4, 0, 0, 2.4))
p_wide <- xs_diagram(wide_bed, water_level = 1.4, "Wide channel (Manning)",
                      "5/3", "Width much greater than depth")

grid.arrange(p_rect, p_vnotch, p_wide, ncol = 3)
```

Read the shapes as top-width behaviour, since that's what actually drives
the exponent: a rectangular section presents the same top width at every
head, so discharge grows only as fast as velocity does, close to
head^1.5. A V-notch's top width grows linearly with head, so discharge
picks up an extra factor of head on top of that, close to head^2.5. A
wide channel sits between the two -- width barely changes as depth
increases a little, so hydraulic radius tracks depth almost directly, and
Manning's equation gives roughly head^(5/3). None of this is a
`reach.rate` model; it's what `rate_optimise()`'s unconstrained fit is
implicitly competing against when gaugings are too sparse to pin the
exponent down on their own.

# Why the shape sets the exponent

$Q = C(H-a)^n$ is a straight line on log-log axes, with slope $n$
(`rating_curves_guide.Rmd` states this; here it is drawn):

```{r loglog-slope, fig.width=6, fig.height=4.5}
H <- seq(0.1, 3, by = 0.02)
C <- 5
curves_dt <- rbindlist(lapply(c(1.5, 5 / 3, 2.5), function(n) {
  data.table(H = H, Q = C * H^n, n = sprintf("n = %s", if (n == 5 / 3) "5/3" else n))
}))

ggplot(curves_dt, aes(H, Q, colour = n)) +
  geom_line(linewidth = 1.1) +
  scale_x_log10() +
  scale_y_log10() +
  scale_colour_manual(values = c("n = 1.5" = "#0288d1", "n = 5/3" = "#01579b", "n = 2.5" = "#7b1fa2")) +
  labs(x = "Head, H - a (m, log scale)", y = "Discharge, Q (m³/s, log scale)", colour = NULL) +
  theme_minimal(base_size = 12)
```

The three lines cross where the curves happen to agree (`H = 1` here,
an artefact of sharing one `C`, not a general property) and fan out
either side of it -- the steeper the line, the larger `n`, exactly
matching which cross-section above is steepest-sided in its top-width
growth. This is the number `n_bounds = c(lower, upper)` constrains
directly: not the curve's height (`C` absorbs that), its steepness.

# A compound channel: the case that actually needs separate limbs

Real stations rarely stay one shape across their whole range. A channel
with a floodplain shelf looks like one control at low flow and a
completely different one once the river tops its banks:

```{r xs-compound, fig.width=6.5, fig.height=4.5}
compound_bed <- data.table(
  distance_m = c(-12, -5, -3, 3, 5, 12),
  elevation_m = c(2.6, 2.0, 0, 0, 2.0, 2.6)
)
baseline <- min(compound_bed$elevation_m) - 0.4
levels_dt <- data.table(
  label = c("In-bank", "Overbank"), elevation_m = c(1.2, 2.4),
  colour = c("#0288d1", "#01579b")
)
wet_inbank <- clip_below(compound_bed, levels_dt$elevation_m[1])
wet_inbank_poly <- rbindlist(list(
  wet_inbank,
  data.table(distance_m = rev(wet_inbank$distance_m), elevation_m = levels_dt$elevation_m[1])
))
wet_overbank <- clip_below(compound_bed, levels_dt$elevation_m[2])
wet_overbank_poly <- rbindlist(list(
  wet_overbank,
  data.table(distance_m = rev(wet_overbank$distance_m), elevation_m = levels_dt$elevation_m[2])
))
ground_poly <- data.table(
  distance_m = c(compound_bed$distance_m, rev(compound_bed$distance_m)),
  elevation_m = c(compound_bed$elevation_m, rep(baseline, nrow(compound_bed)))
)

ggplot() +
  geom_polygon(data = ground_poly, aes(distance_m, elevation_m), fill = bed_colour, colour = NA) +
  geom_polygon(data = wet_overbank_poly, aes(distance_m, elevation_m), fill = levels_dt$colour[2], alpha = 0.45) +
  geom_polygon(data = wet_inbank_poly, aes(distance_m, elevation_m), fill = levels_dt$colour[1], alpha = 0.75) +
  geom_line(data = compound_bed, aes(distance_m, elevation_m), colour = line_colour, linewidth = 1.1) +
  geom_hline(data = levels_dt, aes(yintercept = elevation_m, colour = label), linetype = "dashed", linewidth = 0.6) +
  geom_label(
    data = levels_dt, aes(x = 11, y = elevation_m, label = label, colour = label),
    hjust = 1, vjust = -0.35, size = 3.2, fontface = "bold", label.size = 0.3, fill = "white"
  ) +
  scale_colour_manual(values = setNames(levels_dt$colour, levels_dt$label), guide = "none") +
  labs(
    title = "Compound channel + floodplain",
    subtitle = "n changes once the floodplain engages",
    caption = "Top width jumps at bankfull -- a new limb, a new n, not a bigger n_bounds",
    x = NULL, y = NULL
  ) +
  coord_cartesian(ylim = c(baseline, 3.1)) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 11),
    plot.caption = element_text(size = 8, colour = "grey40", hjust = 0),
    axis.text = element_blank(), axis.ticks = element_blank(),
    panel.grid = element_blank()
  )
```

In-bank, the top width barely changes with stage -- a narrow, near-vertical-
sided channel, closer to the rectangular family. Once stage crosses the
bankfull shoulder, top width jumps abruptly onto the wide floodplain, and
the control the exponent should track jumps with it. This is exactly
`rating_curves_guide.Rmd`'s "Why ratings have limbs" argument, seen from
the geometry side: it is also a real constraint on `n_bounds` itself.
Reading `R/rate_optimise.R`'s fitting loop, `n_bounds` is one
`c(lower, upper)` pair applied to *every* limb inside a single
`rate_optimise(..., control = ...)` call -- there's no per-limb variant.
A compound channel like this one needs two different bounds for its two
regimes, which means two separate `rate_optimise()` calls (one per limb's
data), the same pattern `vignette("rating_curve_walkthrough")`'s top-limb
section already uses -- not one call with a single `n_bounds` wide enough
to straddle both controls, which would fail to constrain either one
usefully.

# Choosing bounds in practice

| Control | Typical literature `n` | A reasonable starting `n_bounds` |
|---|---|---|
| Rectangular weir / channel | ~1.5 | `c(1.4, 1.6)` |
| Wide channel (Manning) | ~5/3 (1.67) | `c(1.5, 1.8)` |
| V-notch / triangular | ~2.5 | `c(2.3, 2.7)` |

A few practical notes that don't fit neatly in the table:

- **Anchor near the literature value, don't pin it.** `c(1.4, 1.6)`
  around a expected 1.5 leaves the fit some room; `c(1.5, 1.5)` forces
  the exponent outright and should be reserved for when you are
  confident enough to fix it, not as a default width.
- **Widen or drop the bound as gaugings accumulate.** `n_bounds` earns
  its keep exactly where `non_standard_optimisation.Rmd` introduces it:
  sparse, high-stage data with nothing else to anchor the shape. Once a
  limb has enough of its own gaugings spanning its range, the
  unconstrained fit is usually trustworthy on its own, and a stale bound
  just risks quietly overriding what the data would otherwise show.
- **A hard constraint, not a prior.** `n_bounds` is a box constraint on
  the optimiser, not a Bayesian prior the data can update -- see
  `non_standard_optimisation.Rmd`'s "What this is not" section (and
  [issue #11](https://github.com/JonPayneEA/reach.rate/issues/11)) for
  the distinction. Use it when you're confident enough in the control
  type to rule other exponents out, not as a softer nudge.
- **Compound geometry needs compound handling**, per the previous
  section -- one `n_bounds` per control, one `rate_optimise()` call per
  control.

# A worked example, tied back to the diagram

`non_standard_optimisation.Rmd`'s own bounds demonstration already uses a
`true_n = 1.5` station -- the rectangular-weir shape from this vignette's
first section, not a new one:

```{r worked-example}
set.seed(23)
stage_dense_m <- seq(0.2, 2.5, by = 0.03)
stage_sparse_m <- c(2.7, 2.9, 3.1, 3.3)
stage_m <- c(stage_dense_m, stage_sparse_m)

true_C <- 5; true_a <- 0; true_n <- 1.5
discharge_true_cms <- true_C * (stage_m - true_a)^true_n
discharge_cms <- discharge_true_cms * exp(rnorm(length(stage_m), sd = 0.1))

fit_unconstrained <- rate_optimise(discharge_cms, stage_m)
fit_bounded <- rate_optimise(discharge_cms, stage_m, n_bounds = c(1.4, 1.6))

rbind(
  data.table(fit = "unconstrained", fit_unconstrained@limbs[, .(C, a, n, rmse_cms, r_squared)]),
  data.table(fit = "n_bounds = c(1.4, 1.6)", fit_bounded@limbs[, .(C, a, n, rmse_cms, r_squared)])
)
```

The unconstrained `n` drifts away from the rectangular control's expected
1.5, because the handful of gaugings above stage 2.5 aren't enough on
their own to hold it there. `n_bounds = c(1.4, 1.6)` -- the row in this
vignette's own table, chosen directly from that shape's typical range --
holds it to the range the diagram at the top of this vignette says a
rectangular control should occupy. `rmse_cms`/`r_squared` barely move:
the data cannot distinguish the two shapes here, which is exactly why the
physical anchor matters.

# Where to go next

- `vignette("non_standard_optimisation")` -- the mechanical
  demonstration of `n_bounds` (and `objective = "relative"`, and the
  `gauging_datetime` groundwork).
- `vignette("rating_curves_guide")` -- "Why a power law" for the full
  derivation and citations, and "Why ratings have limbs" for the
  compound-channel argument in prose.
- `vignette("rating_curve_walkthrough")` -- `n_bounds` used in context,
  anchoring a real sparse top limb against an extreme gauging.

# References

Herschy, R. W. (2009). *Streamflow Measurement* (3rd ed.). Routledge.

Rantz, S. E., et al. (1982). *Measurement and Computation of Streamflow*
(Water Supply Paper 2175). U.S. Geological Survey.
