---
title: "Non-Standard Rating Optimisation: When and Why"
author: "Jonathan Payne, Environment Agency Flood Forecasting"
date: "August 2026"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Non-Standard Rating Optimisation: When and Why}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE, echo = TRUE, collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5
)
library(reach.rate)
library(data.table)
```

# Why deviate from the default fit

`rate_optimise()`'s default call -- no `objective`, no `n_bounds`, no
`gauging_datetime` -- makes three assumptions that are entirely reasonable
most of the time:

1. Every gauging counts equally in *absolute* discharge terms: a 1 m³/s
   miss is a 1 m³/s miss, whether the true flow is 2 m³/s or 2,000 m³/s.
2. The exponent `n` in `Q = C(H-a)^n` is estimated purely from whatever
   gaugings you have -- no outside knowledge of the channel is brought in.
3. Every gauging counts equally regardless of *when* it was taken.

Each of these can fail to hold. This vignette covers the two situations
where the package already has an answer -- proportional gauging error
(`objective = "relative"`) and sparse, physically-constrained extremes
(`n_bounds`) -- and one where the groundwork exists but the fix doesn't yet
(gauging age, `gauging_datetime`). See `vignette("rating_curves_guide")`
for the standard workflow these options extend, and `ROBUST_FITTING.md` in
the package sources for a shorter, code-free version of this same material.

# When absolute error is the wrong error model

## The problem

Streamflow gauging accuracy is conventionally expressed as a *percentage*
of the reading, not a fixed number of cubic metres per second (Herschy
2009) -- a gauging is typically accurate to within a few percent whether
the flow is a trickle or a torrent. The default fit doesn't know this: it
minimises absolute squared error, so a limb with both low and high flows in
it will fit the high flows preferentially, simply because their absolute
residuals are numerically larger. The low-flow shape can suffer as a
result.

## Demonstration

```{r relative-setup}
set.seed(11)
stage_m <- seq(0.2, 3.5, by = 0.03)
true_C <- 4; true_a <- 0.05; true_n <- 1.7
discharge_true_cms <- true_C * (stage_m - true_a)^true_n
# Proportional (percentage) noise, not fixed-magnitude noise -- this is
# what "gauging accuracy is a few percent of the reading" looks like.
discharge_cms <- discharge_true_cms * exp(rnorm(length(stage_m), sd = 0.08))
```

```{r relative-fit}
fit_absolute <- rate_optimise(discharge_cms, stage_m)
fit_relative <- rate_optimise(discharge_cms, stage_m, objective = "relative")

rbind(
  data.table(objective = "absolute", fit_absolute@limbs[, .(C, a, n, rmse_cms, r_squared)]),
  data.table(objective = "relative", fit_relative@limbs[, .(C, a, n, rmse_cms, r_squared)])
)
```

Both rows report `rmse_cms`/`r_squared` on the same natural discharge
scale, so they're directly comparable -- `objective` only changes what the
optimiser treats as "best," not how the result is reported afterward. With
proportional noise like this, `objective = "relative"`'s `C`/`a`/`n` sit
closer to the true generating values, because the low-flow gaugings (whose
proportional errors are just as informative as the high-flow ones'
proportional errors) get to pull their proper share of the fit's weight.

```{r relative-plot}
plot_rating_curves(Absolute = fit_absolute, Relative = fit_relative)
```

## When to reach for it

If your gaugings span a wide flow range and you know (or suspect) accuracy
is roughly proportional across that range -- true of most current-meter and
ADCP gauging -- `objective = "relative"` is a reasonable default choice,
not just a fix for a diagnosed problem. It costs nothing when the two
objectives happen to agree.

# When the top of the curve needs a physical anchor

## The problem

The top of a rating -- the part covering the biggest floods -- is
structurally the least-gauged part of the record, because big floods are
rare and often dangerous or impractical to gauge directly. Left alone, the
fit has to infer the curve's steepness up there from whatever thin evidence
exists, which can drift to an implausible exponent. But open-channel
hydraulics already tells us roughly what to expect: about 1.5 for a
rectangular control, about 2.5 for a triangular/V-notch one, and Manning's
equation gives roughly 5/3 for a wide channel (Herschy 2009; Rantz et al.
1982) -- see `vignette("rating_curves_guide")`'s "Why a power law" section
for the full derivation. `n_bounds` lets that theory anchor the fit instead
of leaving it to guess. For what these controls actually look like, and
how to read a cross-section into a starting `n_bounds`, see
`vignette("n_bounds_guide")`.

## Demonstration

```{r bounds-setup}
set.seed(23)
# A channel control theory says should be close to rectangular (n ~ 1.5),
# but the fit only has a handful of gaugings above stage 2.5 -- exactly
# the "sparse extremes" situation this section is about.
stage_dense_m <- seq(0.2, 2.5, by = 0.03)
stage_sparse_m <- c(2.7, 2.9, 3.1, 3.3)
stage_m <- c(stage_dense_m, stage_sparse_m)

true_C <- 5; true_a <- 0; true_n <- 1.5
discharge_true_cms <- true_C * (stage_m - true_a)^true_n
discharge_cms <- discharge_true_cms * exp(rnorm(length(stage_m), sd = 0.1))
```

```{r bounds-fit}
fit_unconstrained <- rate_optimise(discharge_cms, stage_m)
fit_bounded <- rate_optimise(discharge_cms, stage_m, n_bounds = c(1.4, 1.6))

rbind(
  data.table(fit = "unconstrained", fit_unconstrained@limbs[, .(C, a, n, rmse_cms, r_squared)]),
  data.table(fit = "n_bounds = c(1.4, 1.6)", fit_bounded@limbs[, .(C, a, n, rmse_cms, r_squared)])
)
```

With so little data above stage 2.5, the unconstrained exponent can drift
well away from the true, physically-expected value of 1.5 -- an
extrapolation risk that matters most exactly where a design flood
assessment would use the curve. `n_bounds = c(1.4, 1.6)` holds the fit to
the range channel-control theory actually supports, at the cost of a
(usually small) increase in fitted error over the *sparse, noisy* region it
constrains. `rate_optimise_constrained()` honours the same bound across
every limb it refits, not just the one it leaves alone, so a multi-limb
rating stays consistent with the constraint throughout.

## What this is not

Le Coz et al. (2014) make the general case for combining hydraulic
knowledge with uncertain gaugings when estimating a rating -- `n_bounds` is
this package's answer to that case, but it is a **hard or soft box
constraint**, not a **prior**. A constraint rules a range of `n` in or out
outright; a true Bayesian prior instead expresses a starting belief that
the data can still overturn, given enough of it. That version -- where
channel-control theory would be a prior distribution on `n` rather than a
bound -- is a separate, larger, not-yet-started idea tracked as
[issue #11](https://github.com/JonPayneEA/reach.rate/issues/11). Reach for
`n_bounds` when you're confident enough in the channel-control type to rule
out other exponents; wait for #11 if you'd rather let the data override a
merely plausible assumption.

# When gauging age matters

## The problem

Channels change: the bed shifts, vegetation grows and dies back, sediment
moves, a flood re-cuts the cross-section (Mansanarez et al. 2019, "Shift
happens!", describe exactly this and how to adjust a rating when it
occurs). A rating fitted from gaugings spanning many years implicitly
treats a measurement from a decade ago as equally informative about
*today's* channel as one from last month, which won't always be true.

## What exists today

The fix for this -- weighting gaugings by age, without discounting a
genuinely large but old flood just because it's old -- is not yet built;
see [issue #9](https://github.com/JonPayneEA/reach.rate/issues/9) and the
package's `ROBUST_FITTING.md` for the open design questions (decay shape,
how "large" is defined, per-limb vs. whole-record weighting).

What *has* landed is the data the fix will need. `rate_optimise()` and
`rate_optimise_segmented()` accept an optional `gauging_datetime`:

```{r datetime-demo}
gauging_datetime <- as.Date("2015-01-01") + round(seq_along(stage_dense_m) * 20)
fit_dated <- rate_optimise(
  discharge_cms[seq_along(stage_dense_m)], stage_dense_m,
  gauging_datetime = gauging_datetime
)
head(fit_dated@gaugings)
```

Nothing in the fit reads this column yet -- `fit_dated`'s coefficients are
identical to what an undated call would produce. It's stored so that when
the weighting math in #9 is designed, it has real dates to work with rather
than needing its own follow-up data-model change first.

# Where to go next

`vignette("rating_curves_guide")` covers the standard fitting workflow
these options extend. `ROBUST_FITTING.md` in the package sources gives a
shorter, code-free version of this same material, written for anyone who
wants the reasoning without the R. The GitHub issues linked throughout
(#9, #11) carry the open technical and design questions for what isn't
built yet.

# References

Herschy, R. W. (2009). *Streamflow Measurement* (3rd ed.). Routledge.

Le Coz, J., Renard, B., Bonnifait, L., Branger, F., & Le Boursicaud, R. (2014). Combining hydraulic knowledge and uncertain gaugings in the estimation of hydrometric rating curves: a Bayesian approach. *Journal of Hydrology*, 509, 573-587.

Mansanarez, V., Westerberg, I. K., Lam, N., & Lyon, S. W. (2019). Shift happens! Adjusting stage-discharge rating curves to morphological changes at known times. *Water Resources Research*, 55(8), 6642-6661.

Rantz, S. E., et al. (1982). *Measurement and Computation of Streamflow* (Water Supply Paper 2175). U.S. Geological Survey.
