---
title: "Absolute vs Relative: How the Two Fitting Objectives Actually Work"
author: "Jonathan Payne, Environment Agency Flood Forecasting"
date: "September 2026"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Absolute vs Relative: How the Two Fitting Objectives Actually Work}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE, echo = TRUE, collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5
)
library(reach.rate)
library(data.table)
library(ggplot2)
```

`vignette("non_standard_optimisation")` covers `objective = "relative"`
mechanically: what it does to a fit, and a worked comparison showing its
`C`/`a`/`n` land closer to the truth on proportionally-noisy data. It
doesn't show *why* -- what the optimiser is actually doing differently
underneath. This vignette is that missing diagram.

# Two misses that look very different depending how you measure them

Picture two gaugings, each missed by exactly 1 m³/s: one on a small
stream reading 2 m³/s (the model predicted 3), one on a large river
reading 200 m³/s (the model predicted 201).

```{r two-misses, fig.alt="Two hypothetical gaugings each missed by 1 cubic metre per second, shown as equal absolute bars but wildly different percentage bars"}
misses_dt <- data.table(
  station = factor(c("Small stream (2 m^3/s)", "Large river (200 m^3/s)"),
    levels = c("Small stream (2 m^3/s)", "Large river (200 m^3/s)")
  ),
  absolute_miss = c(1, 1),
  pct_miss = c(50, 0.5)
)

p_abs <- ggplot(misses_dt, aes(station, absolute_miss)) +
  geom_col(fill = "#0288d1", width = 0.5) +
  labs(title = "Absolute objective sees", x = NULL, y = "Miss (m^3/s)") +
  theme_minimal(base_size = 11) + theme(axis.text.x = element_text(size = 8))

p_pct <- ggplot(misses_dt, aes(station, pct_miss)) +
  geom_col(fill = "#e65100", width = 0.5) +
  labs(title = "Relative objective sees", x = NULL, y = "Miss (%)") +
  theme_minimal(base_size = 11) + theme(axis.text.x = element_text(size = 8))

gridExtra::grid.arrange(p_abs, p_pct, ncol = 2)
```

The absolute objective (the package's default, `Q = C(H-a)^n` fitted
directly) sees these two misses as *identical* -- both contribute
`1^2 = 1` to the sum of squared errors, so the optimiser has no more
reason to fix one than the other. But the small-stream miss is a 50%
error and the large-river miss is a 0.5% error -- wildly different by any
practical measure of gauging accuracy, which is conventionally expressed
as a percentage of the reading, not a fixed number of m³/s (Herschy
2009). The relative objective (`log(discharge_cms)` fitted instead) sees
percentages, not absolute misses -- so it treats the small-stream error as
the one worth fixing, a hundred times over.

# The equivalence, made explicit

That difference in what gets penalised is exactly equivalent to an
implied *weight* on each gauging, even though `objective = "relative"`
never mentions the word "weight." Minimising squared log-residuals,

$$\sum_i \big(\log Q_{\text{obs},i} - \log Q_{\text{fit},i}\big)^2,$$

is, for small residuals, approximately the same as minimising *weighted*
squared absolute residuals with weight `1/Q_fit^2` -- because
$\log Q_{\text{obs}} - \log Q_{\text{fit}} \approx (Q_{\text{obs}} -
Q_{\text{fit}})/Q_{\text{fit}}$ (a first-order Taylor expansion, the same
approximation behind treating a lognormal error model as "percentage
error"), so

$$\big(\log Q_{\text{obs}} - \log Q_{\text{fit}}\big)^2 \approx
\frac{1}{Q_{\text{fit}}^2}\big(Q_{\text{obs}} - Q_{\text{fit}}\big)^2.$$

`objective = "relative"` never computes this weight explicitly -- it
fits in log-space directly, which amounts to the same thing without ever
constructing the vector -- but plotting it makes the mechanism visible:

```{r implied-weight, fig.alt="Implied per-gauging weight against stage: flat at 1 for the absolute objective, sharply higher at low flow for the relative objective"}
set.seed(11)
stage_m <- seq(0.2, 3.5, by = 0.03)
true_C <- 4; true_a <- 0.05; true_n <- 1.7
discharge_true_cms <- true_C * (stage_m - true_a)^true_n
discharge_cms <- discharge_true_cms * exp(rnorm(length(stage_m), sd = 0.08))

fit_relative <- rate_optimise(discharge_cms, stage_m, objective = "relative")
fitted_cms <- fit_relative@limbs$C * (stage_m - fit_relative@limbs$a)^fit_relative@limbs$n

weight_dt <- data.table(
  stage_m = rep(stage_m, 2),
  weight = c(rep(1, length(stage_m)), 1 / fitted_cms^2),
  objective = rep(c("absolute", "relative"), each = length(stage_m))
)
weight_dt[, weight_norm := weight / max(weight), by = objective]

ggplot(weight_dt, aes(stage_m, weight_norm, colour = objective)) +
  geom_line(linewidth = 1.1) +
  labs(
    title = "What each objective effectively weights a gauging by",
    subtitle = "Normalised to a maximum of 1, for comparison",
    x = "Stage (m)", y = "Implied weight (normalised)", colour = "objective"
  ) +
  theme_minimal(base_size = 11)
```

Absolute weights every gauging the same, at every flow. Relative weights
low-flow gaugings far more heavily than high-flow ones -- which is
exactly the behaviour `non_standard_optimisation.Rmd` observed
indirectly (the low-flow shape holding up better under `objective =
"relative"`); this is the mechanism producing it.

# Seeing it in the residuals

The same story shows up in how each fit's errors are distributed, if you
look at *percentage* residuals rather than absolute ones -- the scale
gauging accuracy is actually judged on. A raw scatter of percentage
residuals against stage is too noisy at any realistic sample size to
show this by eye; summarising as percentage RMSE within stage bands
makes it clear instead:

```{r residual-comparison, fig.alt="Percentage RMSE by stage band and objective, as a grouped bar chart -- the absolute objective's percentage error is much larger in the lowest stage band, the relative objective's is roughly even across all three"}
fit_absolute <- rate_optimise(discharge_cms, stage_m)
fitted_abs <- fit_absolute@limbs$C * (stage_m - fit_absolute@limbs$a)^fit_absolute@limbs$n

resid_dt <- rbind(
  data.table(stage_m = stage_m, objective = "absolute", fitted_cms = fitted_abs),
  data.table(stage_m = stage_m, objective = "relative", fitted_cms = fitted_cms)
)
resid_dt[, discharge_cms := rep(discharge_cms, 2)]
resid_dt[, pct_residual := 100 * (discharge_cms - fitted_cms) / fitted_cms]
resid_dt[, stage_band := cut(stage_m, 3, labels = c("Low", "Mid", "High"))]

rmse_pct_dt <- resid_dt[, .(rmse_pct = sqrt(mean(pct_residual^2))), by = .(objective, stage_band)]

ggplot(rmse_pct_dt, aes(stage_band, rmse_pct, fill = objective)) +
  geom_col(position = "dodge") +
  labs(
    title = "Percentage RMSE by stage band",
    subtitle = "Absolute objective's error concentrates in the low-stage band; relative's is roughly even",
    x = "Stage band", y = "RMSE (%)", fill = "objective"
  ) +
  theme_minimal(base_size = 11)
```

The absolute fit's percentage error is markedly worse in the low-stage
band than anywhere else -- exactly where percentage errors matter most
and absolute errors matter least, so the optimiser had the least reason
to get it right there. The relative fit's percentage error sits roughly
even across all three bands: no stage band is systematically favoured
over another, because the objective it actually minimised was already
expressed on that same percentage scale.

# When to reach for it

Unchanged from `non_standard_optimisation.Rmd`'s own guidance: if
gaugings span a wide flow range and accuracy is roughly proportional
across it -- true of most current-meter and ADCP gauging -- `objective =
"relative"` is a reasonable default, not just a fix for a diagnosed
problem. What this vignette adds is the mechanism: it isn't a different
kind of curve, just a different, and for most gauging programmes more
realistic, definition of what counts as a big error.

# References

Herschy, R. W. (2009). *Streamflow Measurement* (3rd ed.). Routledge.

# Where to go next

- `vignette("non_standard_optimisation")` -- the mechanical API
  (`objective = "relative"`) and a worked `C`/`a`/`n` recovery
  comparison on proportionally-noisy synthetic data.
- `vignette("rating_curves_guide")` -- "Fitting: `rate_optimise()`" for
  where `objective` sits among the function's other arguments.
- `?rate_optimise` -- full argument reference.
