---
title: "Rating Methods: How the Pieces Fit Together"
author: "Jonathan Payne, Environment Agency Flood Forecasting"
date: "September 2026"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Rating Methods: How the Pieces Fit Together}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE, echo = TRUE, collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5
)
library(reach.rate)
library(data.table)
library(ggplot2)
```

This package started as one function, `rate_optimise()`, and one vignette.
It is now five ways to build a rating, several ways to check one, and half
a dozen ways to use one once you have it -- each with its own vignette
going deep on its own corner. None of them, on its own, shows how the
corners connect. This vignette is that map: two flow diagrams, read
top to bottom, and the reasoning behind each arrow.

Throughout both diagrams: **grey** boxes are the data you start with,
**blue** boxes are functions that fit or bridge between representations,
**green** boxes are the S7 objects you actually hold at each stage,
**amber** boxes check or fix something, and **purple** boxes are what you
do with a finished rating. Nothing here replaces the vignette on any
individual box -- every box names the function whose own documentation
and `@examples` you'd actually run.

# Three ways to start

The first decision is not which function to call -- it's what you have.
A station with a gauging history, a freshly-surveyed cross-section with
no gaugings yet, and a calibrated weir all produce a rating (or a
discharge estimate) by genuinely different routes.

```{r diagram-start, echo=FALSE, fig.alt="Flow diagram: gaugings feed rate_optimise(), rate_optimise_constrained(), or rate_optimise_segmented(), producing a FlodeRating or FlodeSegmentedRating; a surveyed cross-section feeds rate_from_cross_section(), also producing a FlodeRating; weir or flume geometry feeds the weir/flume discharge equations, producing a standalone discharge table rather than a FlodeRating.", fig.width=11.5, fig.height=5.5}
boxes <- data.table(
  id = c("gaug", "xs", "weir",
         "ro", "roc", "ros", "rfcs", "wd",
         "fr", "fseg", "standalone"),
  label = c(
    "Gaugings\n(discharge, stage[, datetime])",
    "Surveyed cross-section\n(distance, elevation)",
    "Weir/flume geometry\n(known dimensions)",
    "rate_optimise()\nindependent limbs",
    "rate_optimise_constrained()\nboundary-matched limbs",
    "rate_optimise_segmented()\njoint multiplicative model",
    "rate_from_cross_section()\nManning's equation",
    "weir_discharge_*()\nflume_discharge_parshall()",
    "FlodeRating", "FlodeSegmentedRating",
    "Discharge + uncertainty\n(data.table, standalone)"
  ),
  x = c(3.3, 9.7, 13.3,
        0, 3.3, 6.6, 9.7, 13.3,
        4.5, 7.6, 13.3),
  y = c(9, 9, 9,
        7, 7, 7, 7, 7,
        5, 5, 5),
  w = c(3.3, 2.8, 2.6,
        2.6, 2.9, 2.9, 2.6, 2.7,
        2.0, 2.9, 2.7),
  h = c(0.9, 0.9, 0.9,
        0.9, 0.9, 0.9, 0.9, 0.9,
        0.8, 0.8, 0.9),
  fill = c(rep("#cfd8dc", 3), rep("#b3e5fc", 5), "#c8e6c9", "#c8e6c9", "#f5f5f5")
)

arrows <- data.table(
  x    = c(3.3, 3.3, 3.3,  9.7,  13.3,   0,   3.3,   9.7,  6.6,   13.3),
  y    = c(8.55, 8.55, 8.55, 8.55, 8.55, 6.55, 6.55, 6.55, 6.55, 6.55),
  xend = c(0, 3.3, 6.6,  9.7,  13.3,   4.5, 4.5, 4.5,   7.6,   13.3),
  yend = c(7.45, 7.45, 7.45, 7.45, 7.45, 5.4, 5.4, 5.4, 5.4, 5.4)
)

ggplot() +
  geom_rect(data = boxes, aes(xmin = x - w/2, xmax = x + w/2, ymin = y - h/2, ymax = y + h/2, fill = fill),
            colour = "grey30") +
  geom_text(data = boxes, aes(x, y, label = label), size = 2.5, lineheight = 0.9) +
  geom_segment(data = arrows, aes(x = x, y = y, xend = xend, yend = yend),
               arrow = arrow(length = unit(0.15, "cm"), type = "closed"), colour = "grey30") +
  annotate("text", x = 13.3, y = 5.85, label = "standalone -- not a FlodeRating,\nno gaugings/fitting involved",
           size = 2.3, colour = "grey40", lineheight = 0.9) +
  scale_fill_identity() +
  coord_cartesian(xlim = c(-1.9, 15.25), ylim = c(4.3, 9.7)) +
  theme_void()
```

**You have gaugings.** This is the default case and the one most of the
package assumes. Three functions read the same `(discharge_cms, stage_m)`
pair, and choosing between them is a modelling decision, not a
preference:

- `rate_optimise()` -- fits each limb independently. Simplest, most
  transparent, but adjacent limbs rarely meet exactly at their shared
  boundary (a "junction gap"), which the next section's diagram picks up.
- `rate_optimise_constrained()` -- the same independent-limb shape, but
  refit so adjacent limbs are forced to agree at each boundary as part of
  the fit itself, not as a separate step afterwards.
- `rate_optimise_segmented()` -- a structurally different joint model
  (Hodson et al. 2024): one multiplicative expression across the whole
  range, so there is no junction to reconcile because there was never a
  junction. See `vignette("segmented_model_guide")` for how the factors
  compose.

**You have a surveyed cross-section but no gaugings.** `rate_from_cross_section()`
derives a theoretical rating from channel geometry alone via Manning's
equation, then hands the resulting synthetic `(stage, discharge)` series
to `rate_optimise()` internally -- which is why its output is an ordinary
`FlodeRating`, distinguished only by `@provenance$source`. Useful at an
ungauged or newly-installed site, or as a physically-grounded sanity
check on what a surveyed section implies about curve shape. See
`vignette("weir_flume_guide")`'s sibling discussion and `?rate_from_cross_section`.

**You have a weir or flume with known dimensions.** The four
`weir_discharge_*()`/`flume_discharge_parshall()` functions compute
discharge directly from the structure's own published equation and
geometry, with GUM-style propagated uncertainty -- no gaugings, no
`rate_optimise()` call, no `FlodeRating` at the end of it. This is the
one box in the diagram that doesn't feed into anything else in this
package: it's a standalone calculator, not a fitting pipeline. In
principle a structure's computed discharge at a given head could be
treated as one very-high-confidence synthetic gauging and blended into a
`rate_optimise()` call the way `rate_from_cross_section()` blends
Manning-derived points -- the package doesn't automate that today; see
`vignette("weir_flume_guide")` for the equations themselves.

# From a fit to a product

Once you hold a `FlodeRating` or `FlodeSegmentedRating`, what happens
next genuinely differs between the two -- not because one is more
finished than the other, but because a joint segmented model has no
junction to check and no natural table form the way independently-fitted
limbs do.

```{r diagram-downstream, echo=FALSE, fig.alt="Flow diagram: FlodeRating branches to diagnostic functions (a dead end) and to as_rating_table(), producing a FlodeRatingTable, which passes through gap detection and resolution to a gap-free FlodeRatingTable, which feeds apply_rating(), apply_rating_inverse(), apply_rating_versioned(), and compare_ratings(). FlodeSegmentedRating goes directly to apply_rating() and rating_plot(), with no gap-check step and no diagnostic-function equivalent, and also bridges via as_rating_table() to a FlodeSegmentedRatingTable, which supports apply_rating() and apply_rating_inverse() but not apply_rating_versioned() or compare_ratings().", fig.width=12, fig.height=9}
boxes2 <- data.table(
  id = c("fr", "fseg", "diag", "art", "segapply", "frt1", "gapcheck", "frt2",
         "app1", "app2", "app3", "app4",
         "art_seg", "fsegtable", "apps_seg"),
  label = c(
    "FlodeRating", "FlodeSegmentedRating",
    "plot_rating_residuals()\nflag_extrapolated_limbs()\nflag_influential_gaugings()\nplot_rating_leverage()",
    "as_rating_table()",
    "apply_rating()\nrating_plot()",
    "FlodeRatingTable",
    "expand_rating_table() ->\ndetect_rc_gaps() -> resolve_rc_gaps() /\nalign_limb_equations() / align_limb_boundaries() /\ngraft_rating()  (only if gaps found)",
    "FlodeRatingTable\n(gap-free)",
    "apply_rating()", "apply_rating_inverse()", "apply_rating_versioned()",
    "compare_ratings()\nplot_rating_comparison()",
    "as_rating_table()",
    "FlodeSegmentedRatingTable",
    "apply_rating()\napply_rating_inverse()"
  ),
  x = c(1.5, 9,
        -2.3, 1.5, 7.3,
        1.5,
        1.5,
        1.5,
        -3.4, -0.6, 2.5, 5.9,
        10.7, 10.7, 10.7),
  y = c(11, 11,
        9, 9, 9,
        7,
        5,
        3,
        0.8, 0.8, 0.8, 0.8,
        9, 7, 5),
  w = c(2.0, 3.2,
        3.4, 2.6, 3.0,
        2.3,
        5.4,
        2.7,
        2.1, 2.5, 2.7, 3.1,
        2.7, 3.9, 3.4),
  h = c(0.8, 0.8,
        1.5, 0.8, 1.0,
        0.8,
        1.5,
        0.9,
        0.8, 0.8, 0.8, 0.9,
        0.8, 0.8, 1.0),
  fill = c("#c8e6c9", "#c8e6c9",
           "#ffe0b2", "#b3e5fc", "#d1c4e9",
           "#c8e6c9",
           "#ffe0b2",
           "#c8e6c9",
           "#d1c4e9", "#d1c4e9", "#d1c4e9", "#d1c4e9",
           "#b3e5fc", "#c8e6c9", "#d1c4e9")
)

arrows2 <- data.table(
  x    = c(1.5, 1.5, 9,   9,   1.5,  1.5,      1.5,   1.5,  1.5,  1.5,  1.5,   10.7,     10.7),
  y    = c(10.6, 10.6, 10.6, 10.6, 8.6,  6.6,     4.25,  2.55, 2.55, 2.55, 2.55,  8.6,      6.6),
  xend = c(-2.3, 1.5, 7.3, 10.7, 1.5,  1.5,      1.5,   -3.4, -0.6,  2.5,  5.9,   10.7,     10.7),
  yend = c(9.75, 9.4, 9.5, 9.4, 7.4,  5.75,     3.4,   1.2,  1.2,  1.2,  1.25,  7.4,      5.5)
)

ggplot() +
  geom_rect(data = boxes2, aes(xmin = x - w/2, xmax = x + w/2, ymin = y - h/2, ymax = y + h/2, fill = fill),
            colour = "grey30") +
  geom_text(data = boxes2, aes(x, y, label = label), size = 2.4, lineheight = 0.9) +
  geom_segment(data = arrows2, aes(x = x, y = y, xend = xend, yend = yend),
               arrow = arrow(length = unit(0.15, "cm"), type = "closed"), colour = "grey30") +
  annotate("text", x = 7.3, y = 7.9, label = "no gap step needed --\nsingle joint model by construction",
           size = 2.2, colour = "grey40", lineheight = 0.9) +
  annotate("text", x = -2.3, y = 7.55, label = "dead end -- diagnostic only,\nno FlodeSegmentedRating equivalent yet",
           size = 2.1, colour = "grey40", lineheight = 0.9) +
  annotate("text", x = 10.7, y = 3.9, label = "no apply_rating_versioned() or\ncompare_ratings() support yet",
           size = 2.1, colour = "grey40", lineheight = 0.9) +
  scale_fill_identity() +
  coord_cartesian(xlim = c(-4.6, 13.3), ylim = c(0.1, 11.6)) +
  theme_void()
```

**The `FlodeRating` path is the fuller one**, because independently-fitted
limbs create two problems a joint model doesn't have: you might want to
know which gaugings are quietly steering the fit (`plot_rating_residuals()`,
`flag_extrapolated_limbs()`, `flag_influential_gaugings()`,
`plot_rating_leverage()` -- see `vignette("leverage_influence_guide")`),
and adjacent limbs may disagree at their shared boundary
(`as_rating_table()` bridges to a `FlodeRatingTable`, then
`expand_rating_table()` and `detect_rc_gaps()` find any junction gap,
closed by whichever of `resolve_rc_gaps()`, `align_limb_equations()`,
`align_limb_boundaries()`, or `graft_rating()` fits the situation --
`rating_curves_guide.Rmd`'s "The junction gap problem" section covers all
four). What comes out the far end is a gap-free `FlodeRatingTable`, which
is what every application function downstream actually reads.

**The `FlodeSegmentedRating` path is shorter today, not incomplete by
design but genuinely narrower.** `apply_rating()` and `rating_plot()`
both have `FlodeSegmentedRating` methods and work directly on the fit --
no bridging step, because there's no junction to check. It also now
bridges via `as_rating_table()` to a `FlodeSegmentedRatingTable`: one
row per segment (`bp`, `n`) plus the shared `C`, structurally different
from `FlodeRatingTable`'s independent per-row equations (see
`vignette("segmented_model_guide")`'s "Bridging to a table" section),
but enough to support `apply_rating()` again from the bridged shape and,
genuinely new, `apply_rating_inverse()` -- solved numerically
(`stats::uniroot()`) rather than in closed form, since a product of
several shifted power terms has no algebraic inverse the way one term
does. What still doesn't carry over: the four diagnostic functions above
(all four check `S7_inherits(fit, FlodeRating)` and error otherwise --
see `?flag_influential_gaugings`), `apply_rating_versioned()`, and
`compare_ratings()`, the latter two because they expect a
`FlodeRatingTable`'s `lower_level`/`upper_level`/`C`/`a`/`n` shape
specifically, not `FlodeSegmentedRatingTable`'s.

**Coefficient uncertainty runs beside this diagram, not through it.**
`bootstrap_to_table()` and `apply_rating_interval()` (or, for a quick
closed-form alternative, the `C_se_asymp`/`a_se_asymp`/`n_se_asymp`
columns `rate_optimise()` always populates) work directly from a
`FlodeRating`'s own `@bootstrap`/refit machinery, independent of whether
the table above is gap-free yet -- `rating_curves_guide.Rmd`'s
"Coefficient uncertainty" section covers that pipeline on its own terms.

# Where to go next

This vignette is deliberately shallow on every individual box -- it
exists to show the shape of the whole, not to replace any one part of
it. Depending on which box you're standing at:

- `vignette("rating_curves_guide")` -- the full introductory walkthrough
  `rate_optimise()` itself, the junction-gap problem, and coefficient
  uncertainty in depth.
- `vignette("segmented_model_guide")` -- how `rate_optimise_segmented()`'s
  multiplicative factors actually compose, with diagrams.
- `vignette("weir_flume_guide")` -- the weir/flume equations and
  `rate_from_cross_section()`'s Manning calculation, with structure
  diagrams.
- `vignette("leverage_influence_guide")` -- which gaugings are quietly
  steering a `FlodeRating` fit, in plain English.
- `vignette("n_bounds_guide")` -- reading a channel cross-section into a
  starting `n_bounds` for `rate_optimise()`/`rate_optimise_constrained()`.
- `vignette("objective_guide")` and `vignette("recency_weighting_guide")`
  -- two ways `rate_optimise()`'s own fit can be steered, beyond
  `n_bounds`: `objective = "relative"`, and `age_halflife`.
- `vignette("s7_objects_guide")` -- what `@`-access, `S7_inherits()`, and
  the `@status`/`@previous` audit chain actually buy you, for every green
  box in both diagrams above.
