---
title: "A Rating Curve Walkthrough: From Spot Gaugings to a Constrained Curve"
author: "Jonathan Payne, Environment Agency Flood Forecasting"
date: "August 2026"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{A Rating Curve Walkthrough: From Spot Gaugings to a Constrained Curve}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE, echo = TRUE, collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5
)
library(reach.rate)
library(data.table)
```

# What this vignette is

`vignette("rating_curves_guide")` is the theory-and-reference guide, and
`vignette("non_standard_optimisation")` explains when and why to reach for
non-default fitting. Neither tells one continuous story. This one does:
a single station, one running set of gaugings, taken from "we have some
spot gaugings" all the way through to a rating a forecaster could actually
trust at the top end. It deliberately covers a *narrower* path than
`system.file("examples", "walkthrough.R", package = "reach.rate")`, which
tours every function in the package -- this instead follows one decision
at a time, in the order you would actually make them: fit something,
notice it's wrong, work out why, and fix it.

# The gaugings

Real stations don't arrive with breakpoints attached -- someone has been
out with a current meter or an ADCP on however many occasions the
programme allowed, and the job is to work out what curve, or curves, that
scatter of points implies. This station has three genuine physical
regimes (a low-flow section control, a mid-flow channel control, and a
high-flow regime where the floodplain starts to convey water), though we
won't pretend to know that yet. True to how real programmes work, the
high-flow gaugings are the sparsest -- big floods are rare and hard to
gauge safely, so there are far fewer of them.

```{r gaugings}
set.seed(77)
stage_lo <- seq(0.3, 1.39, by = 0.025)
stage_mid <- seq(1.42, 2.59, by = 0.03)
stage_hi <- seq(2.65, 3.35, by = 0.06)
stage_m <- c(stage_lo, stage_mid, stage_hi)

true_C <- c(2.5, 3.5, 6.0); true_a <- c(0, 0.15, 0.2); true_n <- c(1.3, 1.6, 1.85)
discharge_true <- c(
  true_C[1] * (stage_lo  - true_a[1])^true_n[1],
  true_C[2] * (stage_mid - true_a[2])^true_n[2],
  true_C[3] * (stage_hi  - true_a[3])^true_n[3]
)
discharge_cms <- discharge_true * exp(rnorm(length(stage_m), sd = 0.02))

length(stage_m)
range(stage_m)
```

`true_C`/`true_a`/`true_n` are shown here only so this vignette can check
its own working later; a real analyst has the gaugings and nothing else.
Ninety-six gaugings, stage 0.3 to 3.31 m, 2% proportional noise --
plausible current-meter accuracy (Herschy 2009).

# A single curve, and why it's wrong

The obvious first move: one power law across the whole range.

```{r single-fit}
fit_single <- rate_optimise(discharge_cms, stage_m)
fit_single@limbs[, .(C, a, n, rmse_cms, r_squared)]
```

`r_squared` above 0.95 looks respectable at a glance -- and that is
exactly the trap. Look at what it took to get there: `n` near 7, `a`
pinned near -2.9. No open-channel control on earth produces an exponent
of 7 (Herschy 2009 and Rantz et al. 1982 both put every common control
between about 1.3 and 2.5), and a zero-flow datum almost 3 m below the
lowest gauged stage is not a real hydraulic feature -- it is the
optimiser bending an implausible shape to paper over a curvature the data
actually has. `r_squared` alone would have let this through:

```{r single-residuals, fig.alt="Residuals against stage for the single-curve fit, showing a systematic bend"}
plot_rating_residuals(fit_single)
```

The residuals are not noise -- they swing from strongly positive at low
stage to negative in the middle and positive again at the top, the
textbook signature of forcing one power law onto more than one physical
regime. `rating_plot()` shows the same thing as a curve rather than a
residual:

```{r single-curve-plot, fig.alt="Fitted single curve against the gauged points, showing visible misfit at low stage"}
rating_plot(fit_single)
```

# Where do the limbs split?

Guessing breakpoints by eye is one option; `suggest_breakpoints()` is a
less subjective one. It fits a two-limb model at every candidate stage,
scores each by AIC against the current best (so it charges for the extra
parameters, not just rewards a lower RSS), and greedily keeps whichever
clears both the AIC test and a minimum RSS improvement.

```{r suggest-breaks}
candidates_dt <- suggest_breakpoints(discharge_cms, stage_m, max_breaks = 2L)
candidates_dt[fit_status == "ok"]
breaks <- suggested_breakpoints_vector(candidates_dt)
breaks
```

1.375 and 2.59 -- close enough to the station's real transitions (which,
recall, we are not supposed to know yet) that this is a reasonable
breakpoint set to fit on. `suggest_breakpoints()` narrows the search; it
does not replace walking the reach and checking these against an actual
berm, bank top, or control feature.

# Fitting the multi-limb rating

```{r multi-fit}
fit_multi <- rate_optimise(discharge_cms, stage_m, control = breaks)
fit_multi@limbs[, .(limb, lower_stage_m, upper_stage_m, C, a, n, rmse_cms, r_squared, n_obs)]
```

Night and day next to the single curve. Limb 1's `rmse_cms` alone dropped
by two orders of magnitude, and every `a`/`n` pair now sits in a range an
engineer could actually defend (compare against `true_a`/`true_n` above --
limbs 1 and 2, both well-gauged, land close to the values that generated
them; limb 3, gauged only twelve times, is close but visibly less
certain, which is exactly what thin evidence should look like).

```{r multi-curve-plot, fig.alt="Fitted three-limb rating curve against the gauged points"}
rating_plot(fit_multi)
```

# Checking the junctions

Three limbs fitted independently rarely meet exactly where they join --
each limb only ever sees its own gaugings, so nothing forces continuity
at the breakpoints.

```{r gap-check}
rating_table <- as_rating_table(fit_multi)
rc_raw_dt <- expand_rating_table(rating_table, step = 0.01)
gaps_dt <- detect_rc_gaps(rc_raw_dt)
gaps_dt[, .(junction, stage_break, q_lower_end, q_upper_start, gap_abs, gap_rel, gap_flagged)]
```

Both junctions are flagged -- a 29% jump at 1.375 m, and a much larger
108% jump at 2.59 m where the thinly-gauged top limb meets limb 2. A
hydrograph crossing either stage today would see discharge lurch
sideways, an artefact of how the limbs were fitted, not anything the
river actually does.

## Closing them: `align_limb_boundaries()`

Rather than patch the discretised table, `align_limb_boundaries()` works
on the equations directly: it looks for the stage at which two adjacent
limbs' fitted curves genuinely cross, and moves the shared boundary
there -- no rescaling, no changed coefficients, just a relocated join. It
takes the `FlodeRating` fit directly (no need to call `as_rating_table()`
first):

```{r align-boundaries}
aligned <- align_limb_boundaries(fit_multi)
aligned@table[, .(lower_level, upper_level, lower_level_original, upper_level_original, boundary_adjusted)]
```

Junction 1 relocated -- from 1.375 down to about 0.85 m. That is a
genuine root of `C1*(H-a1)^n1 = C2*(H-a2)^n2`, so it is mathematically
correct, but the shift is bigger than a cosmetic nudge: it lands well
inside limb 1's own gauged range, worth a second look before it goes
into an operational rating. Junction 2's `boundary_adjusted` is `FALSE`,
and the call above also raised a warning explaining why:

```
Warning message:
align_limb_boundaries(): limb 2 and 3's curves do not cross within
[0.8535, 3.31]; leaving this junction's boundary unchanged.
```

Limb 2 and limb 3's curves simply do not cross anywhere within either
limb's own fitted range -- their shapes differ too much, which tracks
with limb 3 being the thinly-gauged one from the start. `align_limb_boundaries()`
declines to invent a crossing that isn't there rather than silently
producing a nonsense join; `align_limb_equations()`'s `C`-only rescale
(which never moves the boundary) is the safer tool for a junction like
this one. The gap that *did* close is confirmed by re-running the check:

```{r gap-recheck}
rc_aligned_dt <- expand_rating_table(aligned, step = 0.01)
gaps_after_dt <- detect_rc_gaps(rc_aligned_dt)
gaps_after_dt[, .(junction, gap_abs, gap_rel, gap_flagged)]
```

```{r plot-gaps, fig.alt="Before and after comparison of the rating curve at each junction, showing the lower junction closed and the upper junction still open"}
plot_rc_gaps(rc_raw_dt, rc_aligned_dt)
```

Junction 1's gap is gone (down to floating-point noise); junction 2's
108% jump is exactly where it was -- correctly untouched, and exactly the
limb that needs more than a boundary relocation.

# A rare flood gauging, and an anchored top limb

That unresolved junction points straight at limb 3: twelve gaugings
trying to pin down three parameters, over the part of the rating that
matters most for flood forecasting and is hardest to gauge. Suppose a
rare high flow finally gets measured -- a flood peak safely above
anything gauged so far:

```{r extreme-gauging}
extreme_stage <- 4.1
extreme_q <- true_C[3] * (extreme_stage - true_a[3])^true_n[3]
set.seed(999)
extreme_q <- extreme_q * exp(rnorm(1, sd = 0.02))
sprintf("Extreme gauging: stage = %.2f m, discharge = %.1f m^3/s", extreme_stage, extreme_q)
```

One more point should only help. Refit limb 3 alone with it added -- and
watch what an unconstrained fit does with it:

```{r extreme-unconstrained}
top_stage <- c(stage_hi, extreme_stage)
top_discharge <- c(discharge_cms[stage_m >= 2.6], extreme_q)

fit_top_unconstrained <- rate_optimise(top_discharge, top_stage)
fit_top_unconstrained@limbs[, .(C, a, n, r_squared, n_obs)]
```

`r_squared` is *higher* than before (0.99) -- and `n` has been dragged
past 3, `a` past -1.9. One extra point at the far end of a thin limb was
enough to swing the exponent into territory no open-channel control
occupies (Herschy 2009; Rantz et al. 1982). This is the same trap as the
single-curve fit at the start: a good fit statistic hiding an
indefensible shape. `n_bounds` exists for precisely this -- anchoring the
exponent to what the physics of a wide, compound channel/floodplain
control actually allows, rather than whatever the optimiser finds when
the evidence is this thin:

```{r extreme-constrained}
fit_top_constrained <- rate_optimise(top_discharge, top_stage, n_bounds = c(1.5, 2.0))
fit_top_constrained@limbs[, .(C, a, n, r_squared, n_obs, near_bound)]
```

`r_squared` is essentially unchanged (0.991 either way -- the data cannot
tell these two shapes apart), but `n` now sits at a defensible 2.0
instead of an indefensible 3.05, and `near_bound` flags exactly what
happened: the fit is pressed against its ceiling, worth another gauging
before this limb is trusted much further. The bound did not push the fit
anywhere the data disagreed with; it only fenced off the part of
parameter space no real control reaches.

One honest note to close on: refitting limb 3 changes its equation, which
would reopen the very junction just closed above -- in an operational
workflow the next step is exactly the one already covered:
`detect_rc_gaps()`, then `align_limb_boundaries()` or
`align_limb_equations()` again, on the new fit.

# Where to go next

- `vignette("rating_curves_guide")` -- the full theory and reference,
  including the segmented alternative and bootstrap uncertainty this
  walkthrough didn't touch.
- `vignette("non_standard_optimisation")` -- more on `n_bounds`, the
  `objective = "relative"` alternative to least squares in absolute
  discharge, and why gauging age will eventually matter too.
- `system.file("examples", "walkthrough.R", package = "reach.rate")` --
  every function in the package, one section at a time.

# References

Herschy, R. W. (2009). *Streamflow Measurement* (3rd ed.). Routledge.

Rantz, S. E., et al. (1982). *Measurement and Computation of Streamflow*
(Water Supply Paper 2175). U.S. Geological Survey.
