---
title: "Rating Curves: Theory, Functions, and Guidance"
author: "Jonathan Payne, Environment Agency Flood Forecasting"
date: "August 2026"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Rating Curves: Theory, Functions, and Guidance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE, echo = TRUE, collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5
)
library(reach.rate)
library(data.table)
```

# What a rating curve is

A river gauging station measures water level continuously. It does not measure flow. Level (stage) is cheap: a pressure transducer or a float in a stilling well, logging every fifteen minutes for decades. Flow (discharge) is expensive: someone has to stand in or above the river with a current meter or an ADCP and measure it directly. A single discharge measurement is called a gauging, and a typical station accumulates a few dozen to a few hundred of them over its life.

The rating curve is the bridge between the two. It is a fitted relationship that converts the continuous stage record into a continuous discharge record, calibrated against the occasional gaugings. Every flow statistic you have ever used from a gauging station, every QMED, every flood peak in HiFlows, passed through a rating curve on its way to you. The quality of the rating bounds the quality of everything downstream of it.

## The equation

This toolkit fits the standard hydrometric form:

$$Q = C\,(H - a)^{n}$$

where $Q$ is discharge (m³/s), $H$ is stage (m), and $C$, $a$, $n$ are fitted coefficients. The offset $a$ deserves a moment. Stage is measured against an arbitrary local datum, not against the point where flow actually ceases. The quantity $H - a$ converts stage to effective head above the level of zero flow, and it is head, not raw stage, that drives flow through a control. When $H - a \le 0$ the river is below its zero-flow level and $Q = 0$ by definition.

## Why a power law

The power law is not a curve-fitting convenience. It is what open-channel hydraulics predicts. Flow over a rectangular weir goes as head to the power 1.5; over a V-notch, head to the power 2.5; Manning's equation for a wide channel gives discharge roughly as depth to the power 5/3. Every common control produces discharge as head raised to some exponent between about 1.3 and 2.5, with $C$ absorbing the geometry and roughness (Herschy 2009; Rantz et al. 1982). Fitting $Q = C(H-a)^n$ is fitting the family of curves the physics says to expect.

A useful property follows: on log axes the power law is a straight line,

$$\log Q = \log C + n \log(H - a)$$

with slope $n$. This is why hydrometrists have plotted ratings on log-log paper for a century, and why a change of slope on that plot signals a change of physical control.

## A convention

Throughout this toolkit, plots put discharge on the x axis and stage on the y axis. This is the hydrometric convention: the rating is read the way the station uses it, from a measured stage across to a discharge. Every plotting function here follows it, including the cross-section demonstration, so the curves are directly comparable.

# Why ratings have limbs

A single power law describes a single control. Real channels change control as they fill. At low flow the station may be controlled by a gravel riffle or a weir crest immediately downstream (a section control). As stage rises the riffle drowns out and the friction of the channel itself takes over (channel control). Higher still, the river tops its banks and the floodplain begins to convey flow, changing the effective geometry entirely.

Each regime approximates a different power law. The standard response is a multi-limb rating: the stage range is split at breakpoints, and each limb gets its own $C$, $a$, $n$. The breakpoints should correspond to real physical transitions, a berm, a bank top, a bypass channel activating, and choosing them is a hydrological judgement, not a statistical one. Section 5 returns to this.

# The objects

The toolkit represents its results as three S7 classes rather than bare tables. The practical consequence for you as a user is small: you access contents with `@` rather than `$`, and the objects carry more than numbers.

`FlodeRating` is the result of `rate_optimise()`. `@limbs` holds one row per limb (bounds, coefficients, diagnostics), `@gaugings` holds the data it was fitted to, `@bootstrap` and `@fit_starts` hold uncertainty and fitting bookkeeping when produced, `@status` records whether the fit is as fitted or has been amended, and `@provenance` records how it was made.

`FlodeSegmentedRating` is the result of `rate_optimise_segmented()`, a structurally different model covered in section 9.

`FlodeRatingTable` is the equation-table representation used for gap checking and application: `lower_level`, `upper_level`, `C`, `a`, `n` per limb, same names and sign convention as `FlodeRating`'s `@limbs`. It carries `@status` and `@previous`, so an amended table references the exact object it was amended from. This is an audit trail by construction rather than by discipline: you cannot end up holding an aligned table with no record of what it was aligned from.

Three shared generics work across the classes: `rating_plot()` draws whichever fit you hand it, `apply_rating()` converts stage to discharge with whichever representation you hold, and `as_rating_table()` converts a fit to the table form. The class system chooses the right implementation; you use one name.

A note on why the fit remembers its own gaugings. Every diagnostic, residual plot, and bootstrap in this toolkit needs the original data, and a rating separated from its gaugings is a rating you can no longer question. Keeping them together is deliberate.

## A running example

Every chunk from here on operates on one running example: a three-limb rating station, gauged independently limb by limb (as real stations usually are -- you don't gauge the whole range in one session), which is exactly the situation that produces junction gaps. The story continues chunk by chunk down to section 8; sections 9 and 10 introduce their own, separate synthetic data since they use a structurally different model. Every number below is synthetic -- the point is the shape of the workflow, not these particular coefficients.

```{r running-example}
set.seed(42)
stage_m <- seq(0.5, 3.5, by = 0.03)
true_limb <- cut(
  stage_m,
  breaks = c(0.5, 1.6, 2.2, 3.5),
  labels = FALSE, include.lowest = TRUE
)
true_coefs_dt <- data.table(C = c(3, 6, 10), a = c(0, 0.1, 0.2), n = c(1.4, 1.6, 1.8))
discharge_cms <- true_coefs_dt$C[true_limb] *
  (stage_m - true_coefs_dt$a[true_limb])^true_coefs_dt$n[true_limb] +
  rnorm(length(stage_m), sd = 0.05)

length(stage_m)
range(stage_m)
```

# Fitting: `rate_optimise()`

```{r fit}
fit <- rate_optimise(discharge_cms, stage_m, control = c(1.6, 2.2))
fit@limbs[, .(limb, C, a, n, rmse_cms, r_squared, n_obs)]
```

## The method

`rate_optimise()` fits each limb by nonlinear least squares: it finds the $C$, $a$, $n$ minimising the residual sum of squares

$$\mathrm{RSS} = \sum_{i}\left(Q_i - C\,(H_i - a)^n\right)^2$$

over that limb's gaugings, using the Levenberg-Marquardt algorithm (Levenberg 1944; Marquardt 1963) as implemented in `minpack.lm::nlsLM`.

Why not just take logs and fit a straight line by ordinary regression? Because the two minimise different things. Log-space regression minimises relative error, which treats a 10% miss at 2 m³/s and a 10% miss at 200 m³/s as equally bad. Least squares in real space minimises absolute error, so the high-flow gaugings dominate the fit. For flood forecasting, where the cost of error is concentrated at high flows, real-space fitting is usually the right default, but be aware of the choice you are inheriting: your low-flow fit is being traded for your high-flow fit. Neither convention is wrong; they answer different questions.

Note also that $a$ is a fitted parameter, not just log-space regression's problem to avoid: the classical log-log method requires knowing $a$ before you can take the logarithm, which is exactly why it was historically estimated by trial and error. Nonlinear fitting estimates all three together.

## Bounds

The fit is constrained so that $C > 0$, $n > 0$, and $H - a > 0$ for every gauging in the limb. Without these, the optimiser is free to wander into regions where the equation is negative, complex, or undefined, and will occasionally converge there on awkward data. The bounds do not push the fit anywhere; they only fence off territory where no valid rating exists.

## Multiple starting points

Nonlinear least squares is iterative: it starts from a guess and improves it. A poor guess can strand the optimiser in a local minimum, a fit that no small adjustment improves but that a very different set of coefficients would beat. By default (`multi_start = TRUE`), each limb is fitted from several starting combinations: a few plausible exponents, a data-driven estimate from a quick log-log regression, a data-scaled $C$. Each converged result is checked for validity (finite, non-decreasing discharge across the limb) and the best valid result wins. Every attempt is recorded in `@fit_starts`, so you can see whether the starts agreed (reassuring) or scattered (a sign the data constrain the fit weakly).

The `near_bound` column flags a fit whose $n$ or $a$ finished pressed against its lower bound. That is worth a second look: it can mean a degenerate, nearly-flat fit, or an offset being forced against the lowest gauging, either of which suggests the limb or its data need attention.

## Reading the diagnostics

`@limbs` reports, per limb: `rmse_cms`, `r_squared`, `n_obs`, `n_unique_stage`, `mean_error_cms`, `median_abs_error_cms`, `max_abs_error_cms`, and `residual_df`.

Three points of guidance. First, `r_squared` close to 1 is necessary but weak: a limb spanning a large discharge range can post an excellent $R^2$ over a mediocre fit, because the variance being explained is mostly the range itself. Compare limbs against each other; the one sitting noticeably below its neighbours is the signal, not any fixed threshold. Second, `rmse_cms` is in the units of your gaugings, so compare it to the gauging uncertainty you would expect at that site, typically 5 to 10% of flow for a good current-meter gauging (WMO 2010). An RMSE well inside gauging uncertainty is as good as the data can support. Third, `n_obs` of three fits three parameters exactly, with no residual degrees of freedom (`residual_df = 0`) and therefore no ability to say whether the equation form suits the data at all. Treat single-figure `n_obs` as provisional.

# Choosing breakpoints: `suggest_breakpoints()`

```{r suggest-breakpoints}
candidates_dt <- suggest_breakpoints(discharge_cms, stage_m, max_breaks = 2L)
suggested_breakpoints_vector(candidates_dt)
```

## The trap this function avoids

Adding a limb always improves the fit. Three extra parameters will always reduce RSS, whether or not a real change of control exists, in the same way a higher-order polynomial always tracks data more closely. Raw fit improvement therefore cannot tell you whether a breakpoint is justified. Some penalty for complexity is required.

## The method

`suggest_breakpoints()` scores each candidate stage by actually fitting a two-limb rating there with `rate_optimise()` and comparing it against the current best model using the Akaike Information Criterion (Akaike 1974):

$$\mathrm{AIC} = n \ln\!\left(\frac{\mathrm{RSS}}{n}\right) + 2k$$

where $n$ is the number of gaugings and $k$ the number of fitted parameters (three per limb). The first term rewards fit; the second charges for it. A candidate breakpoint is only eligible if the two-limb model's AIC beats the one-limb model's and the relative RSS reduction clears `min_improvement` (default 2%), which screens out breakpoints that are statistically cheap but practically negligible.

Multiple breakpoints are found greedily: the best candidate from round one is fixed, and round two searches for a further breakpoint against that improved baseline, at least `min_gap` away from any breakpoint already chosen. The search stops as soon as a round finds nothing that clears the bar, so asking for `max_breaks = 3` will not manufacture a third breakpoint. The full candidate table is returned, one row per candidate per round, with scores, RSS improvements, observation counts each side, and a status; `suggested_breakpoints_vector()` extracts just the selected stages. One caution when reading the table: scores are only comparable within a round, since each round is scored against a different baseline.

## The judgement that remains yours

A statistically supported breakpoint is evidence that the data bend there. It is not evidence that a berm, a bank top, or a drowning weir exists there. Walk the reach, look at the cross-section, check the breakpoint against surveyed features. The function narrows the search; it does not close it. The reverse also holds: a real control change can hide below the statistical threshold if the exponents either side happen to be similar.

# Diagnosing a fit

## `plot_rating_residuals()`

$R^2$ and RMSE say how good a fit is; residuals say where it is weak. This plot shows observed minus fitted discharge against stage, one panel per limb.

```{r plot-residuals, fig.alt = "Residual (observed minus fitted discharge) against stage, faceted by limb"}
plot_rating_residuals(fit)
```

What you are looking for: a curved trend within a panel means the power law is the wrong shape for that limb, often a sign of a missed breakpoint; a widening fan means the fit degrades toward one end of the limb, commonly the low end near the zero-flow datum; a cluster of same-signed residuals suggests something systematic, such as a batch of gaugings taken during a period when the control had shifted. Scattered, zero-centred residuals across the whole panel are what a sound fit looks like.

## `flag_extrapolated_limbs()`

```{r flag-extrapolated}
fit <- flag_extrapolated_limbs(fit)
fit@limbs[, .(limb, doubtful)]
```

A limb's declared range comes from the breakpoints you chose, not from where the gaugings reached. The top limb of most stations is declared up to some design level while its highest actual gauging sits well below: everything between is extrapolation, however good the fit statistics look. Extrapolating a rating is sometimes unavoidable, particularly for flood forecasting, but it should never be invisible.

This function compares each limb's declared bounds against its gauged support and flags a limb `doubtful` where the unsupported distance exceeds `tol_frac` of the limb's width (default 5%). Separately, if you supply an operational range (the stages your forecasting or reporting actually requires), it flags whether the whole rating fails to reach it. The `doubtful` flag is carried through `as_rating_table()` into the expanded table and the plots, so a weakly supported limb stays visibly weakly supported all the way downstream.

# The junction gap problem

Fit each limb independently and their curves will not meet. Limb 1's equation, evaluated at the shared boundary stage, gives one discharge; limb 2's gives another. The mismatch

$$\Delta Q = C_2(H_b + a_2)^{n_2} - C_1(H_b + a_1)^{n_1}$$

at boundary stage $H_b$ is generally non-zero because nothing in either fit knew the other existed. The result is a rating that jumps: a river rising smoothly through the boundary appears to change flow discontinuously, which is hydraulic nonsense and can inject spurious steps into flow records and forecast inputs.

## Detecting it

```{r detect-gaps}
rating_table <- as_rating_table(fit)
rc_raw_dt <- expand_rating_table(rating_table, step = 0.01)
gaps_dt <- detect_rc_gaps(rc_raw_dt)
gaps_dt
```

`expand_rating_table()` evaluates each limb's equation across its range at a fixed stage step, producing the discretised stage-discharge table many EA ratings are maintained in day to day. `detect_rc_gaps()` then reports one row per junction: the discharge each side, the absolute and relative mismatch, and a flag where the mismatch exceeds `tol_abs` (default 0.5 m³/s) or `tol_rel` (default 2%). It also classifies the junction type, since two limbs can also meet with a stage gap or overlap rather than a shared boundary stage, and a discharge comparison only makes sense at a genuinely shared stage. `plot_rc_gaps()` draws the before and after curves with the flagged junctions labelled.

Set the tolerances against what matters at the site. A 0.5 m³/s step is invisible on a large river and material on a small one; the relative tolerance exists for exactly that reason.

# Three ways to close a gap

The toolkit offers three closure strategies because they make different trade-offs, and the right one depends on what you hold and what you need.

## `resolve_rc_gaps()`: patch the table

Adjusts the discretised table rows either side of each flagged junction. `method = "midpoint"` moves both sides to the midpoint of their estimates; `"match_lower_to_upper"`/`"match_upper_to_lower"` move only the single junction row to defer entirely to one limb, for when one side is clearly better gauged than the other; `"extend_lower_to_upper"`/`"extend_upper_to_lower"` go further and rescale the *whole* deferring limb's curve, not just its junction row, so the reshaping is visible across the limb rather than only at its boundary.

```{r resolve-gaps}
rc_fixed_dt <- resolve_rc_gaps(rc_raw_dt, method = "midpoint")
plot_rc_gaps(rc_raw_dt, rc_fixed_dt)
```

This is fast and needs nothing but the table. Its limitation is fundamental: the equations that produced the table are untouched, so anyone re-expanding the rating, or evaluating the equations directly, sees the gap again. Use it when the table is the product and the equations are not going anywhere.

## `align_limb_equations()`: rescale C

```{r align-equations}
aligned_result <- align_limb_equations(rating_table, anchor_limb = 1L)
aligned_result@status
identical(aligned_result@previous, rating_table)
```

Fixes the equations themselves, using only the equations. Hold one limb fixed (the anchor, by default the lowest, usually the best gauged), then for each neighbouring limb solve for the $C$ that forces an exact match at the junction:

$$C_{\text{new}} = \frac{Q_{\text{target}}}{(H_b - a)^{n}}$$

where $Q_{\text{target}}$ is the already-corrected neighbour's discharge at the junction stage $H_b$. The correction propagates outward from the anchor, so every junction along the chain closes exactly, and $a$ and $n$ (the curve's datum and shape) are untouched. The original $C$ is kept alongside the scale factor and percentage change, and the result is a new `FlodeRatingTable` whose `@previous` references the exact pre-alignment object: an auditable amendment, not a silent overwrite.

The cost is honest: the rescaled $C$ is no longer the least-squares optimum for that limb's own gaugings. You have traded a small amount of within-limb fit for continuity, holding shape fixed. When only the equations survive (a legacy rating, an imported table), this is the best available move.

## `rate_optimise_constrained()`: refit under the constraint

```{r constrained-refit}
fit_constrained <- rate_optimise_constrained(discharge_cms, stage_m, control = c(1.6, 2.2))
fit_constrained@limbs[, .(limb, C, a, n, rmse_cms, r_squared, aligned)]
```

When the raw gaugings are available, the job can be done properly. Rather than adjusting an already-fitted equation, each non-anchor limb is refitted against its own gaugings subject to matching its neighbour exactly at the junction. The constraint is built into the model by reparameterisation:

$$Q = Q_{\text{target}}\left(\frac{H - a}{H_b - a}\right)^{n}$$

At $H = H_b$ the bracket equals 1 and $Q = Q_{\text{target}}$ identically, for any $a$ and $n$. Continuity is therefore guaranteed by construction, and the optimiser is free to choose the $a$ and $n$ that best fit the limb's gaugings within that guarantee; $C$ is recovered afterwards as $Q_{\text{target}} / (H_b - a)^n$. Two parameters compensate instead of one, so this generally fits each limb's data better than the C-only rescale. The propagation logic is the same anchor-outward chain.

One honest caveat, reported rather than hidden: any bootstrap uncertainty computed during the initial unconstrained fit describes that fit, not the constrained one, so it is dropped with a warning rather than carried through as if it still applied. The same discipline applies to the multi-start bookkeeping.

## Choosing between them

Equations only, table is the product: `resolve_rc_gaps()`. Equations only, equations are the product: `align_limb_equations()`. Raw gaugings available: `rate_optimise_constrained()`. The comparison below is worth running once to calibrate your intuition for how much the last two actually differ in practice.

```{r compare-alignment-methods}
rating_constrained_table <- as_rating_table(fit_constrained)
method_comparison <- compare_ratings(aligned_result, rating_constrained_table)
method_comparison$discharge[, .(
  max_abs_diff_cms = max(abs(discharge_diff)),
  max_abs_pct_diff = max(abs(discharge_pct_diff), na.rm = TRUE)
)]
plot_rating_comparison(method_comparison, old_label = "C-only rescale", new_label = "Constrained refit")
```

# A different architecture: `rate_optimise_segmented()`

Everything so far reconciles independent limbs after the fact. Hodson et al. (2024), in the USGS `ratingcurve` package, parameterise the problem so there is nothing to reconcile:

$$Q = C \cdot \max(H - b_1,\, 0)^{n_1} \cdot \prod_{j \ge 2} \bigl(\max(H - b_j,\, 0) + 1\bigr)^{n_j}$$

Here $b_1$ is the stage of zero flow and $b_2, \dots, b_k$ are the interior breakpoints. Read the product term carefully: at and below its own breakpoint, each later factor is $(0 + 1)^{n_j} = 1$, exactly inert. Above its breakpoint it begins multiplying flow up (or down, for $n_j < 0$ behaviour in their general form; here exponents are fitted freely). Every segment therefore contributes across the whole range above its threshold rather than owning a disjoint band, transitions are smooth, and continuity is automatic by construction. There is no junction gap to detect, resolve, or align in this model, because the model cannot produce one.

This section uses its own synthetic gaugings -- three multiplicative segments rather than three independent limbs -- since that is what this model structurally fits:

```{r segmented-fit, fig.alt = "Fitted segmented rating curve against gauged points"}
set.seed(7)
stage_seg_m <- seq(0.3, 3.5, by = 0.03)
q1 <- 4 * pmax(stage_seg_m - 0.1, 0)^1.55
q2 <- (pmax(stage_seg_m - 1.6, 0) + 1)^0.9
q3 <- (pmax(stage_seg_m - 2.4, 0) + 1)^1.1
discharge_seg_cms <- q1 * q2 * q3 * exp(rnorm(length(stage_seg_m), sd = 0.03))

fit_seg <- rate_optimise_segmented(discharge_seg_cms, stage_seg_m, control = c(1.6, 2.4))
fit_seg@coefficients
rating_plot(fit_seg)
apply_rating(fit_seg, data.table(stage = c(0.5, 1.5, 3.0)))
```

What this implementation deliberately does not reproduce: Hodson et al. fit by full Bayesian inference, with priors and a posterior over every parameter. This function fits by nonlinear least squares. The reason their paper uses Bayesian machinery is instructive: this model's loss surface is genuinely non-convex, particularly with free breakpoints, and point optimisation can stick. That is exactly why multi-start fitting matters most here, and why interior breakpoints are held fixed at `control` by default; `estimate_breakpoints = TRUE` frees them, accepting the non-convexity their paper warns about. If you need the full uncertainty treatment, use their package; this function gives you their model structure inside this toolkit's conventions.

# Uncertainty: the bootstrap

```{r bootstrap, fig.alt = "Rating curve with a shaded bootstrap prediction interval"}
fit_boot <- rate_optimise(discharge_cms, stage_m, control = c(1.6, 2.2), n_boot = 200L, boot_seed = 11L)
fit_boot@limbs[, .(limb, C, C_se, n, n_se)]
plot_rating_interval(fit_boot)
```

## The idea

A fitted $C$, $a$, $n$ are point estimates. Had the gauging programme gone slightly differently, a visit missed, a flood caught, the coefficients would differ. The bootstrap (Efron 1979) estimates that sensitivity directly: resample each limb's gaugings with replacement, refit, and repeat a few hundred times. The spread of the refitted coefficients approximates the spread you would see across alternative gauging histories. `rate_optimise(n_boot = )` does this per limb, warm-starting each refit from the point estimate, and reports standard errors and 2.5/50/97.5 percentile columns.

Two engineering details matter for trust. A resample can land, by chance, on too few distinct stage values to identify three parameters; such draws are rejected before fitting, recorded with a reason, and if the successful fraction falls below `min_boot_success` a warning is raised. And every draw is recorded in `@bootstrap`, failures included, because a bootstrap where 40% of draws quietly failed is a different (and less trustworthy) object than one where they all converged.

## What the interval means, and what it does not

`bootstrap_to_table()` bridges the draws to `apply_rating_interval()`, which computes, per stage value, the mean, median, and a prediction interval across draws, plus the geometric standard error

$$\mathrm{GSE} = \exp\!\bigl(\mathrm{sd}(\ln Q_{\text{draws}})\bigr)$$

the same multiplicative summary Hodson et al. (2024) report: a GSE of 1.1 reads as roughly plus or minus 10%. `plot_rating_interval()` draws the band around the curve, in the style of their Figure 1.

Be clear about scope. This bootstrap captures one source of uncertainty: sensitivity to which gaugings you happened to have. It does not include stage measurement error, error in the gaugings themselves, uncertainty in the breakpoints, or uncertainty in whether the power law is the right form at all. Treat the interval as a lower bound on the true uncertainty. Expect it to widen toward the edges of the gauged range, where fewer data constrain the fit; an interval of near-constant width usually means `n_boot` is too small to resolve the pattern, not that the uncertainty is genuinely flat. For a full uncertainty treatment, the Bayesian rating literature is the reference point: BaRatin (Le Coz et al. 2014) and Hodson et al. (2024).

# Using a rating

## `apply_rating()`

A stage record with a `datetime` column, spanning both sides of a rating re-rate later in this section, so it does double duty below:

```{r hydrograph}
hydrograph_dt <- data.table(
  datetime = seq(
    as.POSIXct("2024-06-01 00:00", tz = "UTC"),
    by = "90 day", length.out = 40
  ),
  stage = c(
    seq(0.8, 3.8, length.out = 20),
    seq(3.8, 0.8, length.out = 20)
  )
)

flow_dt <- apply_rating(rating_table, hydrograph_dt, stage_col = "stage", out_col = "discharge_cms")
sum(flow_dt$extrapolated) # stage values above the gauged range
flow_dt[c(1, 10, 20, 30, 40)]
```

The step everything else builds towards: stage record in, discharge record out. Each stage value is assigned to the limb whose band it falls in, and that limb's $Q = C(h - a)^n$ is evaluated. Stage at or below the zero-flow datum returns exactly zero. Stage outside every limb is evaluated by extrapolating the nearest limb's equation and flagged in the `extrapolated` column: a value is still returned, because for forecasting an extrapolated estimate is usually more useful than an NA, but it is never silently indistinguishable from an interpolated one. The flag is the contract; respect it downstream.

## `as_rating_table()`: a straight copy

`rate_optimise()` and the equation-table representation write the same equation with the same sign convention: both are $Q = C(H - a)^n$. So converting a fit to a table is a straight column-and-name copy of `C`/`a`/`n`, with no coefficient transformation -- `as_rating_table()` remains a named, tested method (rather than an inline conversion at the call site) so the two representations stay a single, deliberate seam rather than something re-derived wherever it's needed. `as_rating_table()` also carries the `doubtful` flag through, so support weakness survives the change of representation.

## Ratings change over time: `apply_rating_versioned()`

```{r versioned}
rating_history_dt <- data.table(
  version = c("pre-2025", "post-2025"),
  effective_from = as.POSIXct(c("2024-01-01", "2025-06-01"), tz = "UTC"),
  effective_to = as.POSIXct(c("2025-06-01", NA), tz = "UTC"),
  lower_level = c(0.0, 0.0),
  upper_level = c(3.5, 3.5),
  C = c(4.0, 4.6),
  a = c(0.0, 0.0),
  n = c(1.6, 1.6)
)

versioned_flow_dt <- apply_rating_versioned(hydrograph_dt, rating_history_dt)
versioned_flow_dt[c(1, 10, 20, 30, 40)]
```

Riverbeds erode and deposit, vegetation grows and is cleared, channels are realigned. The stage-discharge relation at a station is not one curve but a sequence of curves, each in force for a period, which is why stations are re-rated and why Mansanarez et al. (2019) titled their paper on the subject "Shift happens!". `apply_rating_versioned()` takes a rating history, one row per version and limb with effective date ranges, and applies to each stage observation whichever version was in force at that observation's own timestamp. Timestamps outside every version's range return NA with a warning rather than borrowing the nearest version. Internally it composes `apply_rating()` once per version rather than reimplementing the calculation.

## Documenting a change: `compare_ratings()`

```{r document-change}
comparison <- compare_ratings(rating_table, aligned_result)
plot_rating_comparison(comparison, old_label = "Unaligned", new_label = "Aligned")
```

When you amend a rating, the question that matters is not whether the coefficients moved but how much the discharge a real hydrograph would report has changed, and over what part of the stage range. `compare_ratings()` answers both: a limb-by-limb coefficient diff where the limb counts match, and always a discharge diff computed by running both ratings across their shared stage range. `plot_rating_comparison()` draws the two curves and the difference against stage. A change that only bites above bankfull is a very different amendment, operationally, from one that shifts the whole record, even if the coefficient tables look similar.

# Guidance: reading a fit, in order

When a fit lands on your desk, look at things in this order. First the residual plot: shape problems announce themselves there before any statistic. Second, `n_obs` and `residual_df` per limb: a limb fitted on a handful of points is a hypothesis, not a rating. Third, `rmse_cms` against the gauging uncertainty you would expect at the site; then `r_squared`, comparatively across limbs rather than against a threshold. Fourth, `near_bound` and the `@fit_starts` record: did the optimiser find one clear answer or many mediocre ones? Fifth, the `doubtful` flags: where is this rating extrapolating, and does your operational range depend on those stages? Sixth, the junctions: `detect_rc_gaps()` before and after whatever closure strategy you chose. Last, if the rating will be used for anything where uncertainty matters, run the bootstrap and look at where the band widens, remembering it is a floor on the true uncertainty, not the whole of it.

None of this replaces the walk along the reach. Every statistical signal in this document is a prompt for a physical question.

# Where to go next

Every function used above has its own roxygen documentation with a fuller `@description` and a runnable `@examples` block -- see `?rate_optimise`, `?flode_classes` (for the class hierarchy: `FlodeRatingBase`, `FlodeRating`, `FlodeSegmentedRating`, `FlodeRatingTable`, and the three shared generics), and so on. `system.file("examples", "walkthrough.R", package = "reach.rate")` is a linear, section-by-section script covering the same ground as this vignette plus the cross-section demonstration (`demo_cross_section_rating()`) and the interactive explorer (`rating_curve_explorer()`), meant to be run a section at a time rather than sourced in one go. The testthat files under `tests/testthat/` in the package sources are a third source of worked examples, each tied to a specific behaviour being checked.

# References

Akaike, H. (1974). A new look at the statistical model identification. *IEEE Transactions on Automatic Control*, 19(6), 716-723.

Efron, B. (1979). Bootstrap methods: another look at the jackknife. *The Annals of Statistics*, 7(1), 1-26.

Herschy, R. W. (2009). *Streamflow Measurement* (3rd ed.). Routledge.

Hodson, T. O., et al. (2024). ratingcurve: A Python package for fitting streamflow rating curves. U.S. Geological Survey.

Le Coz, J., Renard, B., Bonnifait, L., Branger, F., & Le Boursicaud, R. (2014). Combining hydraulic knowledge and uncertain gaugings in the estimation of hydrometric rating curves: a Bayesian approach. *Journal of Hydrology*, 509, 573-587.

Levenberg, K. (1944). A method for the solution of certain non-linear problems in least squares. *Quarterly of Applied Mathematics*, 2(2), 164-168.

Mansanarez, V., Westerberg, I. K., Lam, N., & Lyon, S. W. (2019). Shift happens! Adjusting stage-discharge rating curves to morphological changes at known times. *Water Resources Research*, 55(8), 6642-6661.

Marquardt, D. W. (1963). An algorithm for least-squares estimation of nonlinear parameters. *Journal of the Society for Industrial and Applied Mathematics*, 11(2), 431-441.

Rantz, S. E., et al. (1982). *Measurement and Computation of Streamflow* (Water Supply Paper 2175). U.S. Geological Survey.
