---
title: "Recency Weighting: The Maths Behind age_halflife"
author: "Jonathan Payne, Environment Agency Flood Forecasting"
date: "September 2026"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Recency Weighting: The Maths Behind age_halflife}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE, echo = TRUE, collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5
)
library(reach.rate)
library(data.table)
library(ggplot2)
```

A rating fitted from gaugings spanning many years implicitly treats a
measurement from a decade ago as equally informative about *today's*
channel as one from last month. That's usually wrong: riverbeds erode
and deposit, vegetation grows and is cleared, channels get re-cut by a
flood (Mansanarez et al. 2019, "Shift happens!", covers exactly this).
`vignette("non_standard_optimisation")` flagged this as an open gap --
`gauging_datetime` was captured but nothing read it. `age_halflife`
closes that gap: opt-in recency weighting for `rate_optimise()` and
`rate_optimise_segmented()`.

# The formula

Each gauging's contribution to the fit is scaled by a weight between
`age_min_weight` and 1:

$$w_i = w_{\min} + (1 - w_{\min}) \times 0.5^{\,\text{age}_i / h}$$

where $\text{age}_i$ is that gauging's age in days relative to
`reference_datetime` (clamped at 0, so a gauging dated *after* the
reference simply gets full weight 1 rather than something above 1), and
$h$ is `age_halflife` -- the age, in days, at which a gauging's weight
has decayed exactly halfway from 1 to the floor $w_{\min}$
(`age_min_weight`).

```{r weight-curve, fig.alt="Recency weight against gauging age, showing the half-life point and the floor asymptote", fig.width=7.5}
h <- 365 * 3 # 3-year half-life
w_min <- 0.1
age_days <- seq(0, 365 * 15, by = 30)
weight <- w_min + (1 - w_min) * 0.5^(age_days / h)

ggplot(data.table(age_years = age_days / 365, weight), aes(age_years, weight)) +
  geom_line(colour = "#0288d1", linewidth = 1.1) +
  geom_vline(xintercept = h / 365, linetype = "dashed", colour = "grey40") +
  geom_hline(yintercept = w_min, linetype = "dotted", colour = "grey40") +
  annotate("text", x = h / 365, y = 1, label = "  one half-life", hjust = 0, size = 3.3, colour = "grey30") +
  annotate("text", x = 14, y = w_min, label = "age_min_weight floor  ", hjust = 1, vjust = -0.6, size = 3.3, colour = "grey30") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Weight decays by half every age_halflife, down to a floor",
    subtitle = "3-year half-life, age_min_weight = 0.1",
    x = "Gauging age (years)", y = "Weight"
  ) +
  theme_minimal(base_size = 11)
```

Two properties worth reading directly off that curve. First, the decay
is genuinely exponential, not linear -- a gauging loses half its
remaining "extra" weight (above the floor) every `age_halflife`,
regardless of how old it already is. Second, the curve never reaches
zero: it asymptotes to `age_min_weight`, however old a gauging gets.

## Why there's a floor at all

A naive age-only decay, run out far enough, would eventually treat any
sufficiently old gauging as worth almost nothing -- including a limb's
single highest-flow flood, if that happened to be gauged early in the
record. That's exactly backwards: a rare, informative extreme is not
less true for being old, and a limb's high end is usually the part with
the fewest gaugings to begin with (`vignette("leverage_influence_guide")`
covers what happens when one gauging is disproportionately load-bearing;
losing it to age decay is the same problem from a different direction).
`age_min_weight` is the guard against that: whatever else happens, no
gauging's weight falls below it.

**Be clear about what this floor does and doesn't do.** It stops any
single gauging being weighted to *nothing* -- but it is not
magnitude-aware. An old extreme flood and an old ordinary low-flow
gauging decay along exactly the same curve; the floor protects both
equally, not the flood specifically. Building a scheme that down-weights
by age *only among gaugings of similar size* -- so a big old flood keeps
more of its influence than a routine old gauging does -- is a real
refinement this doesn't attempt.
[Issue #9](https://github.com/JonPayneEA/reach.rate/issues/9) tracks that
fuller version; what's built here is the simpler, defensible partial
answer: nothing is ever fully forgotten.

# A worked example

Four gauging campaigns, seven years apart, at a station whose channel
has genuinely drifted -- each campaign's own `C` has crept up over the
twenty years spanned (a growing channel control, easy to imagine as bed
scour or bank erosion opening up the section over time):

```{r worked-data, fig.alt="Discharge against stage, coloured by gauging campaign, showing four campaigns with progressively higher curves"}
stage_seq <- seq(0.3, 3, by = 0.15)
eras <- data.table(year = c(0, 7, 14, 20), true_C = c(3, 4, 5, 6))
set.seed(3)
gaugings <- rbindlist(lapply(seq_len(nrow(eras)), function(i) {
  data.table(
    campaign = i,
    stage_m = stage_seq,
    discharge_cms = eras$true_C[i] * stage_seq^1.5 * exp(rnorm(length(stage_seq), sd = 0.02)),
    gauging_datetime = as.Date("2000-01-01") + eras$year[i] * 365 + round(seq_along(stage_seq) * 0.5)
  )
}))

ggplot(gaugings, aes(stage_m, discharge_cms, colour = factor(campaign))) +
  geom_point(size = 1.8) +
  labs(title = "Four re-gauging campaigns over 20 years", x = "Stage (m)", y = "Discharge (m^3/s)", colour = "Campaign") +
  theme_minimal(base_size = 11)
```

Fitted flat (every gauging counted equally regardless of when it was
taken) against fitted with three-year recency weighting:

```{r fit-comparison}
fit_flat <- rate_optimise(
  gaugings$discharge_cms, gaugings$stage_m,
  gauging_datetime = gaugings$gauging_datetime, n_bounds = c(1.5, 1.5)
)
fit_recent <- rate_optimise(
  gaugings$discharge_cms, gaugings$stage_m,
  gauging_datetime = gaugings$gauging_datetime, n_bounds = c(1.5, 1.5),
  age_halflife = 365 * 3, age_min_weight = 0.02
)

rbind(
  data.table(fit = "flat (unweighted)", fit_flat@limbs[, .(C, a, n, rmse_cms)]),
  data.table(fit = "age_halflife = 3 years", fit_recent@limbs[, .(C, a, n, rmse_cms)])
)
```

The flat fit lands close to 4.5 -- the simple average of the four
campaigns' true `C` (3, 4, 5, 6). The recency-weighted fit sits
noticeably closer to 6, the most recent campaign's true value, without
discarding the older campaigns outright (their gaugings still carry
`age_min_weight`'s worth of influence, visible in `@gaugings`):

```{r weight-table}
fit_recent@gaugings[, .(age_weight = round(mean(age_weight), 3), n_gaugings = .N),
  by = .(campaign_year = data.table::year(gauging_datetime))
][order(-campaign_year)]
```

If the true channel had been stable over those twenty years instead, both
fits would land in the same place -- recency weighting only pulls a fit
away from the flat answer when the data actually gives it a reason to,
by penalising *disagreement* between old and new gaugings, not merely
the passage of time.

# Using it in practice

```{r api-example, eval=FALSE}
rate_optimise(
  discharge_cms, stage_m,
  gauging_datetime = gauging_datetime, # required for any of this to apply
  age_halflife = 365 * 5, # a 5-year half-life
  age_min_weight = 0.1, # never below 10% weight
  reference_datetime = NULL # default: as of this dataset's own most recent gauging
)
```

`age_halflife = NULL` (the default) fits exactly as before -- supplying
`gauging_datetime` alone changes nothing about the fit; recency
weighting only switches on once `age_halflife` is set. There's no
universally right half-life: a station with a fast-shifting bed (regular
dredging, an actively eroding bank) might reasonably use one or two
years; a stable, well-established control might use ten or more, or skip
weighting entirely. `reference_datetime` defaults to the dataset's own
latest gauging -- pass an explicit value (e.g. today's date) if you want
ages measured forward to when the rating is actually being *used*,
rather than back from its last gauging.

`rate_optimise_segmented()` takes the same three arguments, weighting
its single joint fit the same way. `rate_optimise_constrained()`
forwards them through `...` to its own internal `rate_optimise()` call,
same as `gauging_datetime` itself already does.

# References

Mansanarez, V., Renard, B., Coz, J.L., Lang, M., & Darienzo, M. (2019).
Shift happens! Adjusting stage-discharge rating curves to morphological
changes at known times. *Water Resources Research*, 55(4), 2876-2899.

# Where to go next

- `vignette("non_standard_optimisation")` -- the wider set of situations
  where the default fit's assumptions can fail, of which gauging age was
  one.
- `vignette("leverage_influence_guide")` -- what happens when one
  gauging (old or otherwise) is disproportionately steering a limb's
  fit; the same underlying concern the `age_min_weight` floor guards
  against from a different angle.
- `?rate_optimise`, `?rate_optimise_segmented` -- full argument reference.
