---
title: "Working with reach.rate's S7 Objects"
author: "Jonathan Payne, Environment Agency Flood Forecasting"
date: "September 2026"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Working with reach.rate's S7 Objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE, echo = TRUE, collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5
)
library(reach.rate)
library(data.table)
library(ggplot2)
```

Every other vignette in this package uses `FlodeRating`, `FlodeRatingTable`,
and their relatives without pausing to explain them -- `rating_curves_guide.Rmd`'s
"The objects" section gives them one paragraph before moving on to fitting.
This vignette is that paragraph, expanded: what these objects actually
are, how to inspect and check them, and what their built-in guarantees
buy you that a plain list or data.frame wouldn't.

# Why S7 at all

[S7](https://rconsortium.github.io/S7/) is a typed object system for R,
built jointly by the R Consortium and Posit as a successor to S4 that
keeps S4's real guarantees (typed properties, validators that run on
construction) with a plainer syntax. The practical consequence for you
as a user is small and mostly cosmetic: you read a fitted rating's
contents with `@` instead of `$`. What it buys the package is not
cosmetic -- a `FlodeRating` cannot exist in a broken state, because its
validator runs before you ever get to hold one.

# The class hierarchy

```{r hierarchy-diagram, echo=FALSE, fig.alt="Class hierarchy diagram: FlodeRatingBase (abstract) is the parent of FlodeRating and FlodeSegmentedRating; FlodeRatingTable stands separately", fig.height=4}
boxes <- data.table(
  class = c("FlodeRatingBase\n(abstract)", "FlodeRating", "FlodeSegmentedRating", "FlodeRatingTable"),
  x = c(1.8, 0.7, 2.9, 5.9), y = c(2, 0.7, 0.7, 1.35),
  w = c(2.2, 1.9, 2.3, 2.1), h = c(0.7, 0.7, 0.7, 0.7),
  fill = c("#cfd8dc", "#b3e5fc", "#b3e5fc", "#ffe0b2")
)
arrows <- data.table(
  x = c(0.7, 2.9), y = c(1.05, 1.05), xend = c(1.5, 2.2), yend = c(1.65, 1.65)
)

ggplot() +
  geom_rect(data = boxes, aes(xmin = x - w / 2, xmax = x + w / 2, ymin = y - h / 2, ymax = y + h / 2, fill = fill), colour = "grey30") +
  geom_text(data = boxes, aes(x, y, label = class), size = 3.4, lineheight = 0.9) +
  geom_segment(data = arrows, aes(x = x, y = y, xend = xend, yend = yend),
    arrow = arrow(length = unit(0.2, "cm"), type = "closed"), colour = "grey30") +
  annotate("text", x = 5.9, y = 0.85, label = "separate -- not part of\nthe fit hierarchy", size = 3, colour = "grey40", lineheight = 0.9) +
  scale_fill_identity() +
  coord_cartesian(xlim = c(-0.5, 7.2), ylim = c(0, 2.6)) +
  theme_void()
```

`FlodeRatingBase` is abstract -- you never construct one directly; it
exists only to hold what `FlodeRating` (the result of `rate_optimise()`)
and `FlodeSegmentedRating` (the result of `rate_optimise_segmented()`)
have in common: `@gaugings`, `@fit_starts`, `@status`, `@provenance`.
`FlodeRatingTable` sits apart from that hierarchy entirely -- it's the
equation-table representation `gap_check`'s functions and `apply_rating()`
work with, not a fit in its own right, and it doesn't inherit anything
from `FlodeRatingBase`.

# `@`-access and `print()`

```{r access-basics}
set.seed(1)
stage_m <- seq(0.3, 3, by = 0.1)
discharge_cms <- 5 * stage_m^1.5 * exp(rnorm(length(stage_m), sd = 0.03))
fit <- rate_optimise(discharge_cms, stage_m)

fit # each class has its own print method
class(fit) # the full inheritance chain, most-specific first
fit@limbs[, .(C, a, n, rmse_cms, r_squared)]
```

`print()` gives a one-line summary plus the coefficient table, rather
than R's default (and much less readable) dump of every property.
`class(fit)` shows the full chain -- `reach.rate::FlodeRating`,
`reach.rate::FlodeRatingBase`, `S7_object` -- confirming a `FlodeRating`
really is a `FlodeRatingBase` underneath, matching the diagram above.

# Validators: what you can't accidentally build

The property that matters most in practice: constructing one of these
classes with something structurally wrong fails immediately, with a
specific error, rather than producing an object that only breaks later
when some downstream function trips over the missing piece.

```{r validator-demo, error=TRUE}
FlodeRating(
  limbs = data.table(limb = 1, C = 1, a = 0, n = 1), # missing lower_stage_m/upper_stage_m
  gaugings = data.table(discharge_cms = 1, stage_m = 1)
)
```

That error fires at construction time, naming exactly which columns are
missing -- not three functions later, as an obscure `NULL` or a cryptic
subscript error somewhere that merely *assumed* those columns existed.
Every exported class in this package validates its own structural
requirements (required columns, `C`/`n` positivity, contiguous limb
bounds, and more) the same way.

# `S7_inherits()`: accepting either representation

Several exported functions -- `plot_rating_cross_section()`,
`apply_rating_inverse()`, `graft_rating()` among them -- accept either a
`FlodeRating` or a `FlodeRatingTable` and quietly bridge the former to
the latter internally. `S7_inherits()` is how they tell which one they
were handed, and it's directly usable in your own code if you're writing
a function meant to work with whichever representation the caller holds:

```{r s7-inherits-demo}
S7::S7_inherits(fit, FlodeRating)
S7::S7_inherits(fit, FlodeRatingBase) # also TRUE -- FlodeRating IS a FlodeRatingBase

rating_dt <- data.table(lower_level = 0, upper_level = 3, C = 5, a = 0, n = 1.5)
tbl <- FlodeRatingTable(table = rating_dt)
S7::S7_inherits(tbl, FlodeRating) # FALSE -- a table is not a fit

accept_either <- function(x) {
  if (S7::S7_inherits(x, FlodeRating)) "got a fit" else if (S7::S7_inherits(x, FlodeRatingTable)) "got a table" else stop("neither")
}
accept_either(fit)
accept_either(tbl)
```

`inherits(x, "FlodeRating")` (base R's own generic) does *not* work
reliably here -- S7 classes are namespace-qualified
(`"reach.rate::FlodeRating"`, visible in the `class(fit)` output above),
so `S7_inherits()` exists specifically to check against the class object
itself rather than a string you'd have to get exactly right.

# Reading the audit chain: `@status` and `@previous`

`FlodeRatingTable` (and, via `FlodeRatingBase`, the fit classes too)
carries `@status` and `@previous` so that amending a rating never
silently loses what it was amended from:

```{r audit-chain}
rating_dt <- data.table(
  lower_level = c(0, 1.5), upper_level = c(1.5, 3),
  C = c(3, 5), a = c(0, 0.1), n = c(1.5, 1.6)
)
tbl <- FlodeRatingTable(table = rating_dt)
tbl@status
is.null(tbl@previous)

aligned <- align_limb_equations(tbl)
aligned@status
aligned@previous@status
identical(aligned@previous@table, tbl@table) # the exact object it was built from
```

`aligned@previous` isn't a copy or a re-derived approximation -- it's the
literal pre-amendment object. Chain several amendments and you can walk
the whole history backwards, `@previous@previous@previous...`, until you
reach a `NULL` -- the original, independently-fitted or independently-built
starting point. This is what `vignette("rating_curves_guide")` means when
it calls this "an audit trail by construction rather than by discipline":
there is no code path that produces an amended table with nowhere to
point `@previous` at.

# Shared generics: one name, dispatched by class

Three functions work identically regardless of which class you're
holding, because they're registered as S7 methods against shared
generics rather than being separate functions per class:

```{r shared-generics, eval=FALSE}
rating_plot(fit) # works on FlodeRating, FlodeSegmentedRating, FlodeRatingTable
apply_rating(fit, stage_dt) # same
as_rating_table(fit) # converts any fit class to FlodeRatingTable
```

You don't need to know or check which class you're holding to call any
of these three -- S7's method dispatch picks the registered implementation
for whatever class the object actually is. That's the payoff of the
class hierarchy in the diagram above: it's not just documentation of
what properties exist where, it's what makes "call the same function
regardless of representation" actually work.

# Where to go next

- `vignette("rating_curves_guide")` -- "The objects" for where this
  material first appears in context, and the running example these
  classes are built around throughout the rest of that vignette.
- `?FlodeRatingBase`, `?FlodeRating`, `?FlodeSegmentedRating`,
  `?FlodeRatingTable` -- full property reference for each class.
- [S7's own documentation](https://rconsortium.github.io/S7/) for the
  object system itself, independent of this package.
