---
title: "How the Segmented Model Actually Composes"
author: "Jonathan Payne, Environment Agency Flood Forecasting"
date: "September 2026"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{How the Segmented Model Actually Composes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE, echo = TRUE, collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5
)
library(reach.rate)
library(data.table)
library(ggplot2)
```

`vignette("rating_curves_guide")`'s section on `rate_optimise_segmented()`
gives the formula and states, in prose, that there's no junction gap to
reconcile because the model can't produce one. It doesn't show what that
actually looks like. This vignette is the diagram.

# The formula, read as a product of switches

Hodson et al. (2024)'s segmented model writes discharge as one product
across all $k$ segments, not a curve chosen per stage band:

$$Q = C \, (H - b_1)_+^{\,n_1} \; \prod_{j=2}^{k} \big[(H - b_j)_+ + 1\big]^{n_j}$$

where $(x)_+$ means $\max(x, 0)$ and $b_1 < b_2 < \dots < b_k$ are the
breakpoints. Read each factor for $j \geq 2$ as a switch: at and below
its own breakpoint $b_j$, $(H-b_j)_+ = 0$, so that whole factor equals
$1^{n_j} = 1$ -- multiplying by 1 changes nothing, so a segment below its
own breakpoint is *completely inert*, not just small. Above $b_j$, the
factor climbs above 1 and starts contributing. The first factor is
different only in not having the `+ 1`: it can be exactly 0, which is
what makes it the one that reaches zero flow.

# Watching the switches turn on

Refit the walkthrough example from `rating_curves_guide.Rmd`'s own
segmented section and pull the three factors back apart:

```{r fit-and-decompose, fig.alt="Each of the three multiplicative factors plotted separately against stage, showing factors 2 and 3 sitting flat at 1 until their own breakpoint, then rising"}
set.seed(7)
stage_seg_m <- seq(0.3, 3.5, by = 0.03)
q1 <- 4 * pmax(stage_seg_m - 0.1, 0)^1.55
q2 <- (pmax(stage_seg_m - 1.6, 0) + 1)^0.9
q3 <- (pmax(stage_seg_m - 2.4, 0) + 1)^1.1
discharge_seg_cms <- q1 * q2 * q3 * exp(rnorm(length(stage_seg_m), sd = 0.03))

fit_seg <- rate_optimise_segmented(discharge_seg_cms, stage_seg_m, control = c(1.6, 2.4))
co <- fit_seg@coefficients

stage_seq <- seq(0.2, 3.5, by = 0.01)
factors_dt <- rbind(
  data.table(stage_m = stage_seq, factor = "1: C * (H - b1)+ ^ n1", value = co$C * pmax(stage_seq - co$bp1, 0)^co$n1),
  data.table(stage_m = stage_seq, factor = "2: [(H - b2)+ + 1] ^ n2", value = (pmax(stage_seq - co$bp2, 0) + 1)^co$n2),
  data.table(stage_m = stage_seq, factor = "3: [(H - b3)+ + 1] ^ n3", value = (pmax(stage_seq - co$bp3, 0) + 1)^co$n3)
)

ggplot(factors_dt, aes(stage_m, value)) +
  geom_line(colour = "#0288d1", linewidth = 1.1) +
  geom_vline(data = data.table(factor = unique(factors_dt$factor), bp = c(co$bp1, co$bp2, co$bp3)),
    aes(xintercept = bp), linetype = "dashed", colour = "grey40") +
  facet_wrap(~factor, scales = "free_y", ncol = 1) +
  labs(
    title = "The three factors, plotted separately",
    subtitle = "Dashed line marks each factor's own breakpoint",
    x = "Stage (m)", y = "Factor value"
  ) +
  theme_minimal(base_size = 11)
```

Factors 2 and 3 are exactly flat at 1 up to their own dashed breakpoint
line, then start climbing -- precisely the "inert until switched on"
behaviour the formula states, made visible. Factor 1 is the only one
that reaches 0, at its own breakpoint (`b1`, close to the true zero-flow
stage). Multiply all three together at any stage and you get the fitted
discharge; below `b2`, that product is just factor 1 alone (the other
two are inert at 1), so the curve genuinely behaves like a single
one-segment rating down there -- the transition at `b2` isn't a join
between two different models, it's the moment a second factor stops
being inert.

# Why there's no junction gap to reconcile

Fit the same data two ways -- once as independent limbs
(`rate_optimise()`, `vignette("rating_curves_guide")`'s usual junction-gap-
prone approach), once as the segmented joint model -- and overlay both:

```{r gap-comparison, fig.alt="Independent-limb fit shown with a visible discontinuity at each breakpoint, segmented fit shown smooth through the same points"}
fit_independent <- rate_optimise(discharge_seg_cms, stage_seg_m, control = c(1.6, 2.4))

curve_independent <- rbindlist(lapply(seq_len(nrow(fit_independent@limbs)), function(i) {
  L <- fit_independent@limbs[i]
  h <- seq(L$lower_stage_m, L$upper_stage_m, length.out = 100)
  data.table(stage_m = h, discharge_cms = L$C * (h - L$a)^L$n)
}))
curve_segmented <- data.table(
  stage_m = stage_seq,
  discharge_cms = co$C * pmax(stage_seq - co$bp1, 0)^co$n1 *
    (pmax(stage_seq - co$bp2, 0) + 1)^co$n2 * (pmax(stage_seq - co$bp3, 0) + 1)^co$n3
)

ggplot() +
  geom_point(data = data.table(stage_m = stage_seg_m, discharge_cms = discharge_seg_cms),
    aes(stage_m, discharge_cms), colour = "grey70", size = 0.8) +
  geom_line(data = curve_independent, aes(stage_m, discharge_cms, colour = "Independent limbs"), linewidth = 1) +
  geom_line(data = curve_segmented, aes(stage_m, discharge_cms, colour = "Segmented (joint)"), linewidth = 1) +
  geom_vline(xintercept = c(1.6, 2.4), linetype = "dotted", colour = "grey50") +
  coord_cartesian(xlim = c(1.3, 2.7), ylim = c(4, 35)) +
  labs(
    title = "Same data, two architectures, zoomed on both breakpoints",
    x = "Stage (m)", y = "Discharge (m^3/s)", colour = NULL
  ) +
  theme_minimal(base_size = 11)
```

Zoomed on the two breakpoints, the independent-limb curve visibly steps
at each dotted line -- each limb was fitted with no knowledge the other
existed, so nothing forced them to agree at the junction
(`vignette("rating_curves_guide")`'s "The junction gap problem" section
covers this in general, and the tools for closing it after the fact).
The segmented curve passes through both breakpoints without a kink,
because it was never two curves to begin with -- continuity was never at
risk of being violated, since there was only ever one product to
evaluate.

# The trade-off this buys

Nothing here is free. The segmented model's factors for $j \geq 2$ are
parameterised relative to *every* earlier breakpoint at once (each factor
multiplies onto everything below it), so its loss surface is genuinely
non-convex in a way independent per-limb fitting isn't -- exactly why
`rate_optimise_segmented()` defaults to `multi_start = TRUE` and holds
interior breakpoints fixed unless you explicitly ask for
`estimate_breakpoints = TRUE`. The guaranteed continuity comes from
giving up the independence that made each limb's own fit a comparatively
easy, well-behaved three-parameter problem.

# Bridging to a table

Until recently, that was where a `FlodeSegmentedRating`'s pipeline
stopped: `apply_rating()` and `rating_plot()` worked directly on the
fit, but `as_rating_table()` -- the bridge every `FlodeRating` uses to
reach `apply_rating_inverse()`, `apply_rating_versioned()`, and
`compare_ratings()` -- had no method for it. `vignette("rating_methods_overview")`'s
second diagram called this out as a dead end.

It no longer fully is one. `as_rating_table()` now bridges a
`FlodeSegmentedRating` to a [FlodeSegmentedRatingTable]: one row per
segment (`bp`, `n`), plus the shared `C` every segment's factor
multiplies into -- the table-shaped analogue of the formula above, not
a reinterpretation of it:

```{r bridge-to-table}
seg_table <- as_rating_table(fit_seg)
seg_table
```

`apply_rating()` works on this table exactly as it does on `fit_seg`
itself (same formula, evaluated from the bridged shape). The genuinely
new capability is `apply_rating_inverse()`: unlike a single-limb
`FlodeRatingTable`, there's no algebraic inverse for a product of
several shifted power terms, so it solves for stage numerically
(`stats::uniroot()`) per queried discharge, using that the product is
still strictly monotonic above `b1` (every segment's `n > 0`) to
guarantee the root is unique:

```{r bridge-inverse}
forward_dt <- apply_rating(seg_table, data.table(stage = c(1.0, 1.8, 2.6, 3.2)))
forward_dt

apply_rating_inverse(seg_table, forward_dt[, .(discharge)], out_col = "stage_back")
```

What still doesn't carry over: `plot_rating_residuals()`,
`flag_extrapolated_limbs()`, `flag_influential_gaugings()`, and
`plot_rating_leverage()` all require a `FlodeRating` specifically (they
`S7_inherits()`-check for it), and `compare_ratings()` still expects a
`FlodeRatingTable`'s `lower_level`/`upper_level`/`C`/`a`/`n` shape, not
`FlodeSegmentedRatingTable`'s. Those remain open gaps, not solved by
this bridge.

# References

Hodson, T.O., et al. (2024). Rating curve estimation under model and
parameter uncertainty. *Water Resources Research* (segmented power-law
formulation this function implements).

# Where to go next

- `vignette("rating_curves_guide")` -- "A different architecture:
  `rate_optimise_segmented()`" for the full API and what it deliberately
  doesn't reproduce from Hodson et al.'s own Bayesian treatment; "The
  junction gap problem" for the independent-limb side of this
  comparison in full.
- `vignette("rating_methods_overview")` -- the two flow diagrams this
  vignette's "Bridging to a table" section updates one box of.
- `?rate_optimise_segmented` -- full argument reference.
- `?apply_rating_inverse` -- the numeric-inverse details for
  `FlodeSegmentedRatingTable`.
