---
title: "Weir and Flume Equations: Structures, GUM Uncertainty, and Worked Examples"
author: "Jonathan Payne, Environment Agency Flood Forecasting"
date: "September 2026"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Weir and Flume Equations: Structures, GUM Uncertainty, and Worked Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE, echo = TRUE, collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5
)
library(reach.rate)
library(data.table)
library(ggplot2)
library(gridExtra)
library(grid)
```

Everything else in this package fits a rating to gaugings: `rate_optimise()`
estimates `C`/`a`/`n` by nonlinear least squares, and every uncertainty
figure it produces -- bootstrap or asymptotic -- comes from that fit. A
weir or flume is a different kind of rating entirely. Its stage-discharge
relationship comes from the structure's own geometry and a published
discharge coefficient, not from a curve fitted to measurements at this
particular site. There is nothing to fit and no gaugings required --
which also means there is no `rmse_cms` or `r_squared` standing between
you and a wrong transcription of the standard. This vignette exists
because that correctness burden falls entirely on getting the equation
right the first time.

`weir_discharge_rectangular()`, `weir_discharge_vnotch()`,
`weir_discharge_cipoletti()`, and `flume_discharge_parshall()`
(`R/weir_equations.R`) each take a head (or head series) and the
structure's geometry, and return a discharge together with a GUM-style
propagated uncertainty. This vignette covers what each structure looks
like, where its equation comes from, and how the uncertainty is built --
read `?weir_discharge_rectangular` and its siblings for the full
argument reference.

# The four structures

```{r xs-helpers, echo=FALSE}
bed_colour <- "#8d6e63"
water_colour <- "#0288d1"
line_colour <- "#4e342e"

# Same construction as vignette("n_bounds_guide")'s xs_diagram(): a solid
# outline plus a shaded wetted region up to one water level. Reused here
# for a weir plate's own front-elevation profile, not a channel bed --
# the geometry-drawing logic is identical either way.
clip_below <- function(bed_dt, level) {
  n <- nrow(bed_dt)
  pts <- vector("list", 0)
  wet <- bed_dt$elevation_m <= level
  if (wet[1]) pts[[length(pts) + 1]] <- c(bed_dt$distance_m[1], bed_dt$elevation_m[1])
  for (i in seq_len(n - 1)) {
    x1 <- bed_dt$distance_m[i]; y1 <- bed_dt$elevation_m[i]
    x2 <- bed_dt$distance_m[i + 1]; y2 <- bed_dt$elevation_m[i + 1]
    if (wet[i] != wet[i + 1]) {
      xc <- if (x1 == x2) x1 else x1 + (level - y1) / (y2 - y1) * (x2 - x1)
      pts[[length(pts) + 1]] <- c(xc, level)
    }
    if (wet[i + 1]) pts[[length(pts) + 1]] <- c(x2, y2)
  }
  setNames(as.data.table(do.call(rbind, pts)), c("distance_m", "elevation_m"))
}

weir_diagram <- function(bed_dt, water_level, title, note, baseline_drop = 0.3) {
  baseline <- min(bed_dt$elevation_m) - baseline_drop
  wet_dt <- clip_below(bed_dt, water_level)
  wet_poly <- rbindlist(list(
    wet_dt,
    data.table(distance_m = rev(wet_dt$distance_m), elevation_m = water_level)
  ))
  ground_poly <- data.table(
    distance_m = c(bed_dt$distance_m, rev(bed_dt$distance_m)),
    elevation_m = c(bed_dt$elevation_m, rep(baseline, nrow(bed_dt)))
  )

  ggplot() +
    geom_polygon(data = ground_poly, aes(distance_m, elevation_m), fill = bed_colour, colour = NA) +
    geom_polygon(data = wet_poly, aes(distance_m, elevation_m), fill = water_colour, alpha = 0.55) +
    geom_line(data = bed_dt, aes(distance_m, elevation_m), colour = line_colour, linewidth = 1.1) +
    geom_hline(yintercept = water_level, colour = water_colour, linetype = "dashed", linewidth = 0.6) +
    labs(title = title, caption = note, x = NULL, y = NULL) +
    coord_cartesian(ylim = c(baseline, max(bed_dt$elevation_m) + 0.3)) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", size = 12),
      plot.caption = element_text(size = 8, colour = "grey40", hjust = 0),
      axis.text = element_blank(), axis.ticks = element_blank(),
      panel.grid = element_blank()
    )
}
```

Three of the four are front-elevation views: looking straight at the
plate from downstream, the same way `vignette("n_bounds_guide")` draws an
idealised channel cross-section (this reuses that vignette's own diagram
helper -- a weir plate's profile is geometrically the same kind of shape
as a channel bed). Schematic, not to scale.

```{r xs-three, fig.width=9.5, fig.height=3.6, fig.alt="Front-elevation diagrams of a rectangular, V-notch, and Cipoletti weir plate, each with water shaded up to a head above the crest"}
rect_bed <- data.table(distance_m = c(-2.5, -2.5, 2.5, 2.5), elevation_m = c(1.3, 0.5, 0.5, 1.3))
p_rect <- weir_diagram(rect_bed, water_level = 0.5 + 0.2, "Rectangular (Rehbock)",
                        "Full channel width; crest at height P")

vnotch_bed <- data.table(distance_m = c(-1.6, 0, 1.6), elevation_m = c(1.3, 0, 1.3))
p_vnotch <- weir_diagram(vnotch_bed, water_level = 0.35, "V-notch (Kindsvater-Shen)",
                          "Notch angle theta at the apex", baseline_drop = 0.15)

cip_bed <- data.table(distance_m = c(-1.5, -0.5, 0.5, 1.5), elevation_m = c(1.1, 0.6, 0.6, 1.1))
p_cip <- weir_diagram(cip_bed, water_level = 0.6 + 0.2, "Cipoletti (trapezoidal)",
                       "Sides sloped 1 horizontal to 4 vertical")

grid.arrange(p_rect, p_vnotch, p_cip, ncol = 3)
```

A Parshall flume isn't a plate in the flow at all -- it's a shaped
constriction in the channel itself, so it needs a plan view (looking
down on the channel) rather than a front elevation:

```{r flume-plan, echo=FALSE, fig.width=6.5, fig.height=3.2, fig.alt="Plan view of a Parshall flume: converging walls into a parallel throat section, then diverging walls, with the upstream head gauging point Ha marked in the converging section"}
flume_wall_upper <- data.table(
  x = c(0, 3, 5, 7),
  y = c(2.5, 1.4, 1.4, 2.2)
)
flume_wall_lower <- data.table(
  x = flume_wall_upper$x, y = -flume_wall_upper$y
)
ha_x <- 1.2

ggplot() +
  geom_ribbon(
    data = merge(flume_wall_upper, flume_wall_lower, by = "x", suffixes = c("_upper", "_lower")),
    aes(x = x, ymin = y_lower, ymax = y_upper), fill = water_colour, alpha = 0.35
  ) +
  geom_line(data = flume_wall_upper, aes(x, y), colour = line_colour, linewidth = 1.1) +
  geom_line(data = flume_wall_lower, aes(x, y), colour = line_colour, linewidth = 1.1) +
  geom_vline(xintercept = ha_x, colour = "grey30", linetype = "dashed", linewidth = 0.5) +
  annotate("label", x = ha_x, y = 2.7, label = "H[a]~gauging~point", parse = TRUE, size = 3, label.size = 0) +
  annotate("label", x = 4, y = 0, label = "W~(throat~width)", parse = TRUE, size = 3, label.size = 0) +
  labs(title = "Parshall flume (plan view)", caption = "Converging -> throat -> diverging; flow left to right", x = NULL, y = NULL) +
  coord_fixed(ylim = c(-3, 3.2)) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    plot.caption = element_text(size = 8, colour = "grey40", hjust = 0),
    axis.text = element_blank(), axis.ticks = element_blank(),
    panel.grid = element_blank()
  )
```

# Rectangular (Rehbock)

A plate spanning the full width of its approach channel -- no side
contractions -- with a sharp horizontal crest at height $P$ above the
bed. Head $H$ is measured above the crest, upstream of the drawdown
curve. The Rehbock equation (the form ISO 1438:2017 specifies for this
structure) lets the discharge coefficient vary with the ratio of head to
weir height:

$$Q = \frac{2}{3} C_d \sqrt{2g} \, B H^{3/2}, \qquad C_d = 0.602 + 0.083\frac{H}{P}$$

ISO 1438 restricts this to $P < 1\,\mathrm{m}$; `weir_height_m >= 1`
still runs, but warns, since $C_d$'s fit is unvalidated past that point.

```{r rect-example}
weir_discharge_rectangular(
  head_m = c(0.1, 0.2, 0.3), width_m = 1.5, weir_height_m = 0.5,
  u_cd = 0.02, u_head_m = 0.003
)
```

# V-notch (Kindsvater-Shen)

A triangular notch cut into a plate, apex down. Because the flow width
at the water surface grows linearly with head rather than staying fixed,
a V-notch measures small flows far more precisely than a rectangular
weir of similar size -- the reason it's the usual choice for low-flow
gauging. ISO, ASTM, and USBR all specify the Kindsvater-Shen form:

$$Q = \frac{8}{15}\sqrt{2g}\, C_e \tan(\theta/2)\, H_e^{5/2}, \qquad H_e = H + k_h$$

$C_e$ and the small head correction $k_h$ are both fitted polynomials in
the notch angle $\theta$ (degrees); `weir_discharge_vnotch()` evaluates
them exactly as published (Shen 1981), converting $k_h$ from the
published feet to metres only after evaluating the polynomial, not
before -- refitting a unit-converted polynomial from scratch would risk
a transcription error the original doesn't have. Valid for $\theta$
between 25 and 100 degrees; outside that, the function errors rather
than extrapolating an unvalidated fit.

```{r vnotch-example}
weir_discharge_vnotch(
  head_m = c(0.05, 0.1, 0.15), notch_angle_deg = 90,
  u_cd = 0.02, u_head_m = 0.002
)
```

# Cipoletti (trapezoidal)

A trapezoidal notch, sides sloped 1 horizontal to 4 vertical. That slope
isn't arbitrary: it's chosen so the extra discharge through the widening
sides compensates for the end-contraction losses a rectangular weir
suffers, letting the whole structure use one fixed coefficient with no
weir-height correction:

$$Q = 1.859\, L\, H^{3/2}$$

(SI form, derived here from the USBR *Water Measurement Manual*'s
published US customary form, $Q_{\mathrm{cfs}} = 3.367\, L_{\mathrm{ft}}
H_{\mathrm{ft}}^{3/2}$, by direct unit conversion -- shown as a comment
in `R/weir_equations.R` alongside the SI constant, so the derivation is
auditable rather than just asserted.) That simplicity costs accuracy: USBR
describes a well-maintained Cipoletti installation's overall accuracy as
only around $\pm 5\%$, markedly worse than the other three structures
here.

```{r cip-example}
weir_discharge_cipoletti(
  head_m = c(0.1, 0.2, 0.3), crest_length_m = 1.0,
  u_cd = 0.05, u_head_m = 0.003
)
```

# Parshall flume

Not a plate at all -- a shaped constriction built into the channel bed
and walls, converging to a parallel-sided throat and then diverging back
out. The head is gauged upstream in the converging section ($H_a$), and
because the constriction itself (not an obstruction across the whole
flow) creates the head-discharge relationship, a Parshall flume passes
sediment and debris far better than any of the three plate weirs above --
the usual reason to choose one for a station with a mobile bed.
`flume_discharge_parshall()` implements the general free-flow equation
for standard 1-8 ft (0.3048-2.4384 m) throat widths:

$$Q = 2.340\, W^{1.026}\, H_a^{1.522}$$

(again SI-derived from the published US customary form, $Q_{\mathrm{cfs}}
= 4\, W_{\mathrm{ft}}^{1.026} H_{a,\mathrm{ft}}^{1.522}$). Standard sizes
outside that throat-width range use their own individually tabulated
coefficients (USBR ch. 8 §10; ISO 9826), not reproduced here --
`throat_width_m` outside the range warns rather than silently applying
an unconfirmed extrapolation.

```{r parshall-example}
flume_discharge_parshall(
  head_m = c(0.15, 0.3, 0.45), throat_width_m = 0.3048,
  u_cd = 0.03, u_head_m = 0.005
)
```

A Parshall flume only measures correctly in free (unsubmerged) flow.
Pass the downstream head as `hb_head_m` and this checks the standard
0.7 submergence limit for this throat-width range, erroring rather than
applying an uncorrected free-flow equation to submerged conditions:

```{r parshall-submerged, error=TRUE}
flume_discharge_parshall(
  head_m = 0.3, throat_width_m = 0.3048, u_cd = 0.03, u_head_m = 0.005,
  hb_head_m = 0.25
)
```

# GUM-style uncertainty

ISO 1438 expresses a weir's uncertainty analytically rather than
empirically: no gaugings to bootstrap, no residuals to compute an
asymptotic standard error from (`rate_optimise()`'s two ways of getting
an uncertainty, both covered in `vignette("rating_curves_guide")`) --
just the discharge coefficient's own uncertainty and the head
measurement's uncertainty, propagated through the equation by GUM
(JCGM 100:2008) quadrature. For any of these structures, $Q \propto C\,
H_e^{\,n}$, so the combined relative standard uncertainty is

$$\frac{u_c(Q)}{Q} = \sqrt{u(C)^2 + \left(n\, \frac{u(H)}{H_e}\right)^2}$$

where $n$ is that equation's own head exponent (`3/2` for the rectangular
and Cipoletti weirs, `5/2` for the V-notch, `1.522` for the Parshall
flume) and $u(H)$ is the head measurement's absolute standard
uncertainty. Every function here takes `u_cd` and `u_head_m` directly --
required arguments, not defaulted, since a fabricated default would look
authoritative without being traceable to anything. The expanded
uncertainty reported alongside it is `U = coverage_k * u_c` (`coverage_k
= 2` by default, the conventional GUM choice for approximately 95%
coverage under a normal approximation).

Two things follow directly from that formula, worth internalising rather
than just taking on faith:

```{r gum-shape, fig.alt="Relative combined uncertainty against head for a Cipoletti weir, decreasing as head increases", fig.width=7.5, fig.height=4}
heads <- seq(0.03, 0.5, by = 0.01)
out <- weir_discharge_cipoletti(head_m = heads, crest_length_m = 1, u_cd = 0.05, u_head_m = 0.003)

ggplot(out, aes(head_m, u_c_rel)) +
  geom_line(colour = water_colour, linewidth = 1.1) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Relative uncertainty falls as head rises",
    subtitle = "Cipoletti weir, u_cd = 5%, u_head_m = 3 mm (fixed absolute uncertainty)",
    x = "Head (m)", y = "Combined relative uncertainty"
  ) +
  theme_minimal(base_size = 11)
```

First, a fixed absolute head-measurement uncertainty (a staff gauge read
to the nearest few millimetres, say) matters far more at a low head than
a high one -- `u_head_m` is fixed in the plot above, but `u_c_rel` falls
by more than half as head rises from 3 cm to 50 cm. This is the same
shape as `rate_optimise()`'s own diagnostics degrading toward the low
end of a limb (`vignette("rating_curves_guide")`'s "Guidance: reading a
fit, in order" section) -- a structural weir doesn't escape that, it just
gets there analytically instead of empirically.

Second, `u_cd` sets a floor `u_c_rel` never falls below, however precisely
the head is measured -- as `u(H)/H_e -> 0`, the combined uncertainty
converges on `u_cd` itself. A weir's peak accuracy is bounded by how well
its discharge coefficient is actually known, not by instrumentation.

## Choosing `u_cd`

None of the following are defaults this package applies -- they're
starting points from the sources this vignette cites, for a well-built,
correctly-installed, well-maintained structure. A rough field
installation, dimensional tolerance outside the standard's own limits,
or a debris-affected approach channel all push the real figure higher.

| Structure | Illustrative `u_cd` | Source |
|---|---|---|
| Rectangular (Rehbock) | ~1-2% | ISO 1438's own worked uncertainty examples |
| V-notch (Kindsvater-Shen) | ~1-2% | Common literature reference point for a well-installed notch |
| Cipoletti | Overall accuracy ~5% (not a pure `u_cd`) | USBR *Water Measurement Manual*, ch. 7 |
| Parshall flume | ~2% (lab) to ~3-5% (field) | Ferreira et al. (2021), Monte Carlo propagation |

The Cipoletti figure is worth flagging specifically: USBR states it as
an *overall* accuracy, not decomposed into coefficient and head
components the way ISO 1438 structures its own budget -- treat it as
context for the scale `u_cd` needs to be at, not a value to plug in
unexamined.

# Where this fits in `reach.rate`

These four functions are deliberately outside the `FlodeRating`/
`FlodeRatingTable` machinery: there's no NLS fit, no gaugings, no
`provenance` to record beyond the structure's own geometry. They compose
with the rest of the package only informally for now -- a weir-controlled
low-flow limb sitting below a fitted upper limb is a real station
configuration, and `graft_rating()` already joins two independently-sourced
ratings at the stage where they cross (`vignette("rating_curves_guide")`'s
"Extending into ungauged ground" section covers it for a cross-section-
or fitting-derived limb). Wiring a weir equation's own `C`/`a`/`n` into
that same join -- expressing, say, the rectangular case as a one-row
`FlodeRatingTable` limb -- isn't built yet; the three structures with an
exact fixed-form power law (V-notch, Cipoletti, Parshall) would convert
directly, while the rectangular case's head-dependent $C_d$ would need
the same synthetic-points-then-fit approach `rate_from_cross_section()`
already uses for Manning's equation.

# References

Ferreira, H. M. et al. (2021). Uncertainty analysis of discharge
measurements using Parshall flumes. *ScienceDirect*, S2665-9174(21)000714.

ISO 1438:2017. *Hydrometry -- Open channel flow measurement using
thin-plate weirs*.

JCGM 100:2008. *Evaluation of measurement data -- Guide to the expression
of uncertainty in measurement (GUM)*.

Kindsvater, C. E., & Carter, R. W. (1957). Discharge characteristics of
rectangular thin-plate weirs. *Journal of the Hydraulics Division*, ASCE.

Shen, J. (1981). *Discharge Characteristics of Triangular-Notch Thin-Plate
Weirs* (Water-Supply Paper 1617-B). U.S. Geological Survey.

U.S. Bureau of Reclamation (1997). *Water Measurement Manual* (3rd ed.),
ch. 7 (Weirs) and ch. 8 (Flumes).

# Where to go next

- `vignette("rating_curves_guide")` -- "Extending into ungauged ground"
  for `graft_rating()`, the natural next step for a weir-controlled low
  limb joined to a fitted upper one.
- `vignette("n_bounds_guide")` -- the same front-elevation diagram style
  applied to idealised channel controls rather than manufactured
  structures, and where a rating's exponent comes from physically.
- `?weir_discharge_rectangular`, `?weir_discharge_vnotch`,
  `?weir_discharge_cipoletti`, `?flume_discharge_parshall` -- full
  argument reference and citations for each structure.
